---
title: "Chapter 6: Graphical Representation of Causal Effects"
format:
  html: default
  revealjs:
    output-file: 06-graphical-representation-slides.html
  pdf:
    output-file: 06-graphical-representation-handout.pdf
  docx:
    output-file: 06-graphical-representation.docx
---

{{< include ../latex-macros/macros.qmd >}}

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# Nodes named in `conditioned` are drawn as squares, the book's box for conditioning.
draw_dag <- function(spec, conditioned = character()) {
  ggdag::ggdag(
    dagitty::dagitty(spec),
    node = FALSE, text = FALSE, stylized = TRUE
  ) +
    ggdag::theme_dag() +
    ggdag::geom_dag_point(
      ggplot2::aes(shape = name %in% conditioned),
      color = "steelblue", size = 14
    ) +
    ggplot2::scale_shape_manual(
      values = c("FALSE" = 16, "TRUE" = 15),
      guide = "none"
    ) +
    ggdag::geom_dag_text(color = "white", size = 4)
}
```

Causal inference needs expert knowledge and untestable assumptions about the causal network linking treatment, outcome, and other variables.
In the simple settings of Chapters 1-5 that network could stay implicit;
in realistic settings we must state what we know and what we assume.

This chapter introduces causal diagrams, a graphical tool for encoding qualitative causal assumptions.

::: {.callout-tip title="Draw Your Assumptions Before Your Conclusions"}
Put the causal structure you are willing to assume into a diagram before you analyze the data.
A diagram makes the assumptions explicit,
which clarifies conceptual problems and helps investigators communicate [@hernan2020causal, p. 75].
:::

::: {.notes}
This chapter is based on @hernan2020causal [Chapter 6, pp. 75-89].
Chapters 6-9 of the book use causal diagrams to conceptualize problems.
:::

## 6.1 Causal Diagrams (pp. 75-77)

---

```{r}
#| label: fig-dag-6-1
#| echo: false
#| fig-width: 4
#| fig-height: 2.5
#| fig-cap: "Book Figure 6.1: disease severity $L$, heart transplant $A$, death $Y$."
draw_dag('dag {
  L [pos="0,0"]
  A [pos="1,0"]
  Y [pos="2,0"]
  L -> A
  L -> Y
  A -> Y
}')
```

::: {#def-dag}
## Directed Acyclic Graph

A **directed acyclic graph (DAG)** is a set of nodes joined by arrows (directed edges)
in which no node can be reached from itself by following arrows in their direction.
:::

::: {#exm-dag-6-1}
## Book Figure 6.1 Is a DAG

@fig-dag-6-1 has nodes $L$, $A$, $Y$ and arrows $L \to A$, $L \to Y$, $A \to Y$.
The diagrams in this chapter put earlier variables to the left,
so every arrow points rightward and following arrows never returns to a node already visited.
Adding an arrow $Y \to L$ would create the cycle $L \to A \to Y \to L$,
in which $L$ would be a cause of itself,
and the graph would no longer be a DAG.
:::

---

::: {#def-dag-relatives}
## Parents, Descendants, and Ancestors

In a DAG:

- the **parents** of a node $V$ are the nodes with an arrow pointing directly into $V$;
- a node $W$ is a **descendant** of $V$, and $V$ is an **ancestor** of $W$,
  if $W$ can be reached from $V$ by following one or more arrows in their direction;
- the **non-descendants** of $V$ are the nodes other than $V$ that are not descendants of $V$.
:::

::: {#exm-dag-relatives-6-1}
## Parents and Descendants in Book Figure 6.1

In @fig-dag-6-1, $L$ has no parents, the only parent of $A$ is $L$, and the parents of $Y$ are $L$ and $A$.
$A$ and $Y$ are descendants of $L$,
$Y$ is the only descendant of $A$,
and the only non-descendant of $A$ is its parent $L$.
:::

---

::: {#def-causal-dag}
## Causal DAG

A **causal DAG** is a DAG (@def-dag) whose nodes are random variables and in which:

1. an arrow $V \to W$ means that $V$ has a direct causal effect on $W$ for at least one individual in the population,
   where "direct" means not mediated by the other variables on the graph;
2. the absence of an arrow from $V$ to $W$ means that $V$ has no direct causal effect on $W$ for any individual;
3. every common cause of any pair of variables on the graph, measured or not, is also on the graph;
4. every variable is a cause of its descendants (@def-dag-relatives).

These conditions combine the book's description on p. 75 with its formal definition in Technical Point 6.1 [@hernan2020causal, pp. 75-76].
:::

::: {#exm-causal-dag-6-1}
## Reading Book Figure 6.1 as a Causal DAG

In @fig-dag-6-1, $L$ is disease severity, $A$ is heart transplant, and $Y$ is death.
The arrow $L \to A$ says that severity affects whether at least one individual receives a transplant,
and the arrow $A \to Y$ says that transplant directly affects at least one individual's death.
Drawing the graph as a causal DAG also asserts that $L$ is the only common cause of any two of these variables:
for instance, no unmeasured factor affects both transplant and death.
:::

::: {.callout-warning title="What an Arrow Does Not Say"}
A standard causal diagram does not say whether an effect is harmful or protective.
When a variable has two causes, as $Y$ does in @fig-dag-6-1,
the diagram does not say how they interact [@hernan2020causal, p. 75].
:::

::: {.notes}
The modern theory of causal diagrams arose in computer science and artificial intelligence;
see Pearl (2009) and Spirtes, Glymour, and Scheines (2000).
:::

---

### Directed Acyclic Graphs and the Causal Markov Assumption

::: {#def-markov-dag}
## Markov Property

The joint distribution of the variables on a DAG $G$ is **Markov with respect to $G$** if,
conditional on its parents, each variable is independent of its non-descendants (@def-dag-relatives).
:::

::: {#exm-markov-chain}
## The Markov Property for a Chain

Take the DAG $V_1 \to V_2 \to V_3$.
$V_1$ has no non-descendants,
and the only non-descendant of $V_2$ is its parent $V_1$,
so these two variables add no requirement.
The non-descendants of $V_3$ are $V_1$ and its parent $V_2$,
so the Markov property requires exactly $V_3 \ind V_1 \mid V_2$.
:::

---

::: {#def-causal-markov}
## Causal Markov Assumption

The **causal Markov assumption** is that the joint distribution of the variables on a causal DAG (@def-causal-dag) is Markov with respect to that DAG (@def-markov-dag):
conditional on its direct causes (its parents),
each variable is independent of every variable that it does not cause (its non-descendants)
[@hernan2020causal, pp. 76-77].
:::

::: {#exm-causal-markov-lighter}
## Lighters and Lung Cancer

Suppose smoking $L$ causes both carrying a lighter $A$ and lung cancer $Y$,
and carrying a lighter does not cause lung cancer,
so the causal DAG is $A \leftarrow L \to Y$.
$Y$ is a non-descendant of $A$, whose only parent is $L$,
so the causal Markov assumption gives $A \ind Y \mid L$:
among people with the same smoking status, carrying a lighter tells us nothing more about lung cancer.
:::

::: {#rem-common-causes-on-graph}
## Why Every Common Cause Must Be on the Graph

Condition 3 of @def-causal-dag is what makes the causal Markov assumption credible.
In @exm-causal-markov-lighter, leave smoking $L$ off the graph.
$A$ and $Y$ then have no parents and no arrow between them,
so the Markov property would require $A \ind Y$.
But carrying a lighter and lung cancer share the cause smoking,
so they are generically (for all but special parameter values) associated,
and the distribution is then not Markov with respect to the graph without $L$.
:::

---

::: {#prp-markov-factorization}
## Markov Factorization

Let $G$ be a DAG with nodes $V = (V_1, \ldots, V_M)$,
numbered so that no node is an ancestor of a node with a smaller number
(such a numbering exists because $G$ has no cycles).
Suppose $V$ is discrete with probability mass function $f$,
and $f(v) > 0$ for every $v$ in the product of the variables' ranges.
Write $PA_j$ for the parents of $V_j$ and $pa_j$ for their values in $v$.
Then the distribution of $V$ is Markov with respect to $G$ (@def-markov-dag) if and only if, for every $v$,

$$
f(v) = \prod_{j=1}^{M} f(v_j \mid pa_j).
$$ {#eq-markov-factorization}
:::

::: {.proof}
Suppose first that the distribution is Markov with respect to $G$.
By the chain rule, $f(v) = \prod_{j=1}^{M} f(v_j \mid v_1, \ldots, v_{j-1})$.
A descendant of $V_j$ has $V_j$ as an ancestor, so it has a larger number than $V_j$;
hence $V_1, \ldots, V_{j-1}$ are all non-descendants of $V_j$.
They include every parent of $V_j$, because a parent is an ancestor and so has a smaller number.
The Markov property then gives $f(v_j \mid v_1, \ldots, v_{j-1}) = f(v_j \mid pa_j)$,
which is @eq-markov-factorization.

Conversely, suppose @eq-markov-factorization holds, and fix $j$.
Let $D$ be the set of descendants of $V_j$ and $N$ its set of non-descendants.
No variable in $N$ has a parent in $D$ or equal to $V_j$, since it would then be a descendant of $V_j$.
Sum @eq-markov-factorization over the values of the variables in $D$, one at a time,
starting with the one with the largest number.
Each variable summed out at that point has all its children in $D$ with larger numbers, already summed out,
so it appears only in its own factor, which sums to 1.
What remains is

$$
f(v_j, n) = f(v_j \mid pa_j) \prod_{k : V_k \in N} f(v_k \mid pa_k),
$$

where $n$ is the value of $N$ in $v$ and the product does not involve $v_j$.
Summing over $v_j$ gives $f(n) = \prod_{k : V_k \in N} f(v_k \mid pa_k)$,
so $f(v_j \mid n) = f(v_j \mid pa_j)$.
The parents of $V_j$ are in $N$, so this conditional distribution depends on $n$ only through $pa_j$:
$V_j$ is independent of its non-descendants given its parents.
:::

::: {#exm-markov-factorization}
## Markov Factorization Implies an Independence

Take the lighter DAG $A \leftarrow L \to Y$ of @exm-causal-markov-lighter.
The parents are: none for $L$; $L$ for $A$; $L$ for $Y$.
Start from the chain rule, which holds for any distribution, and apply @prp-markov-factorization:

$$
\begin{aligned}
f(l, a, y)
  &= f(l)\, f(a \mid l)\, f(y \mid l, a)
  && \text{(chain rule)} \\
  &= f(l)\, f(a \mid l)\, f(y \mid l)
  && \text{(the only parent of } Y \text{ is } L\text{)}.
\end{aligned}
$$

Dividing both sides by $f(l)$:

$$
\begin{aligned}
f(a, y \mid l)
  &= \frac{f(l, a, y)}{f(l)}
  && \text{(definition of conditional probability)} \\
  &= f(a \mid l)\, f(y \mid l)
  && \text{(substitute the factorization)},
\end{aligned}
$$

so $A \ind Y \mid L$: the missing arrow from $A$ to $Y$ implies a conditional independence.
:::

::: {.callout-note title="Technical Point 6.1: Causal DAGs, Formally"}
The book's Technical Point 6.1 states formally the definitions above:
DAGs, parents, descendants, and ancestors (@def-dag, @def-dag-relatives),
causal DAGs (@def-causal-dag),
the Markov property and its factorization (@def-markov-dag, @prp-markov-factorization),
and the causal Markov assumption (@def-causal-markov).
A causal DAG on its own makes no claim about the distribution of the data;
the causal Markov assumption is what links the two [@hernan2020causal, Technical Point 6.1, p. 76].
:::

---

### Randomized Experiments and Observational Studies

```{r}
#| label: fig-dag-6-2
#| echo: false
#| fig-width: 3
#| fig-height: 1.5
#| fig-cap: "Book Figure 6.2: a marginally randomized experiment."
draw_dag('dag {
  A [pos="0,0"]
  Y [pos="1,0"]
  A -> Y
}')
```

::: {#exm-randomized-dags}
## Randomized Experiments as Causal DAGs

- If transplant is randomized with a probability that depends on severity $L$,
  then $L$ is a common cause of $A$ and $Y$ and must appear (condition 3 of @def-causal-dag):
  @fig-dag-6-1 depicts this conditionally randomized experiment.
- If everyone has the same probability of transplant, $L$ is not a common cause and can be omitted:
  @fig-dag-6-2 depicts this marginally randomized experiment.
- @fig-dag-6-1 can also depict an observational study in which the only parent of $A$ is $L$,
  so that no other cause of $Y$ affects $A$.
:::

::: {.notes}
Chapter 7 shows that being willing to draw @fig-dag-6-1 for an observational study
is the graphical translation of conditional exchangeability $Y^a \ind A \mid L$ for all $a$
[@hernan2020causal, p. 77].
:::

---

### Graphs and Counterfactuals

Each causal DAG has an underlying counterfactual model (Technical Points 6.2 and 6.3 below).
That model justifies the intuitive graphical rules in this chapter,
but conventional diagrams do not show the counterfactual variables.

Single World Intervention Graphs (SWIGs) (Richardson and Robins 2013) put counterfactuals on the graph;
they are introduced in Chapter 7.

::: {.notes}
Causal diagrams encode both causation (our subject-matter knowledge)
and the associations that the causal structure implies.
That simultaneous representation is what makes them attractive [@hernan2020causal, p. 77].
The book's treatment is informal, aiming at conceptual insight rather than rigor.
:::

---

::: {.callout-note title="Technical Point 6.2: Nonparametric Structural Equation Models"}
In the book, every causal DAG stands for a counterfactual model.
The basic one is defined next.
Robins (1986) introduced it as a finest causally interpreted structural tree graph (FCISTG) "as detailed as the data",
and Pearl (2009) showed how to represent it with a DAG.
For exposition the book assumes that every variable can be intervened on,
although its statistical methods do not require this [@hernan2020causal, Technical Point 6.2, p. 77].
:::

::: {#def-npsem}
## Nonparametric Structural Equation Model

Let $G$ be a DAG with nodes $V_1, \ldots, V_M$, numbered as in @prp-markov-factorization.
A **nonparametric structural equation model (NPSEM)** represented by $G$ assumes that there are unobserved random variables (errors) $\epsilon_1, \ldots, \epsilon_M$
and unknown deterministic functions $f_1, \ldots, f_M$ such that,
for each $m$ and each value $pa_m$ of the parents of $V_m$,
the **one-step-ahead counterfactual** $V_m^{pa_m}$, the value $V_m$ would take if its parents were set to $pa_m$,
equals $f_m(pa_m, \epsilon_m)$.
For a node with no parents, such as $V_1$, this reads $V_m = f_m(\epsilon_m)$.

Counterfactuals are assumed to exist for interventions on every variable,
and every factual and counterfactual value is obtained by applying these equations recursively in the numbered order.
:::

::: {#exm-npsem-6-1}
## The NPSEM for Book Figure 6.1

For @fig-dag-6-1 the structural equations (@def-npsem) are $L = f_L(\epsilon_L)$, $A = f_A(L, \epsilon_A)$, $Y = f_Y(L, A, \epsilon_Y)$.
Setting $A$ to $a$ leaves $L$ unchanged, because $L$ is not a descendant of $A$,
so the counterfactual outcome is obtained by substituting $a$ for $A$:
$Y^a = f_Y(L, a, \epsilon_Y)$.
:::

---

::: {.callout-note title="Technical Point 6.3: Which Independences?"}
An FCISTG alone does not imply the causal Markov assumption (@def-causal-markov):
assumptions about the joint distribution of the errors, or of the counterfactuals, must be added.
Two such models are defined next [@hernan2020causal, Technical Point 6.3, p. 78].
:::

::: {#def-npsem-ie}
## NPSEM with Independent Errors

An **NPSEM-IE** is an NPSEM (@def-npsem) whose errors $\epsilon_1, \ldots, \epsilon_M$ are mutually independent.
This is the model that Pearl usually assumes.
:::

::: {#def-ffrcistg}
## FFRCISTG

A **finest fully randomized causally interpreted structured tree graph (FFRCISTG)** is an FCISTG,
that is, a model as in @def-npsem,
in which, for every value $v = (v_1, \ldots, v_M)$ of the nodes,
the one-step-ahead counterfactuals $V_1^{pa_1}, \ldots, V_M^{pa_M}$,
with each $pa_m$ read off from $v$,
are mutually independent (Robins 1986).
:::

::: {#exm-fig-6-2-two-models}
## Two Models for Book Figure 6.2

For @fig-dag-6-2 the structural equations (@def-npsem) are $A = f_A(\epsilon_A)$ and $Y^a = f_Y(a, \epsilon_Y)$.

- As an NPSEM-IE (@def-npsem-ie), $\epsilon_A \ind \epsilon_Y$.
  Then $A$, a function of $\epsilon_A$, is independent of $(Y^{a=0}, Y^{a=1})$, a function of $\epsilon_Y$:
  full exchangeability $(Y^{a=0}, Y^{a=1}) \ind A$ holds.
- As an FFRCISTG (@def-ffrcistg), the one-step-ahead counterfactuals for $v = (a, y)$ are $A$ and $Y^a$,
  so the model gives $Y^a \ind A$ for each $a$ separately:
  marginal exchangeability, which says nothing about $A$ and the pair $(Y^{a=0}, Y^{a=1})$ jointly.
:::

---

::: {#prp-npsem-ie-ffrcistg}
## Every NPSEM-IE Is an FFRCISTG, Not Conversely

Every NPSEM-IE (@def-npsem-ie) is an FFRCISTG (@def-ffrcistg).
The converse fails: some FFRCISTGs are not NPSEM-IEs.
:::

::: {.proof}
Fix $v$.
In an NPSEM-IE, each one-step-ahead counterfactual $V_m^{pa_m} = f_m(pa_m, \epsilon_m)$, with $pa_m$ fixed by $v$, is a function of $\epsilon_m$ alone.
Functions of mutually independent random variables are mutually independent,
so these counterfactuals are mutually independent, as an FFRCISTG requires.

For the converse, take @fig-dag-6-2 with $Y^{a=0}$ and $Y^{a=1}$ independent fair coin flips,
and let $A = 1$ if exactly one of them equals 1 and $A = 0$ otherwise.
Setting $\epsilon_Y = (Y^{a=0}, Y^{a=1})$ and $\epsilon_A = A$ gives an NPSEM.
For each $a$, whatever the value of $Y^a$, the other, independent fair coin makes $A$ equal 1 with probability $1/2$,
so $A \ind Y^a$ and the model is an FFRCISTG.
But $A$ is a function of $(Y^{a=0}, Y^{a=1})$ and is not constant, so full exchangeability fails.
By @exm-fig-6-2-two-models, no NPSEM-IE for @fig-dag-6-2 gives this joint distribution of $(A, Y^{a=0}, Y^{a=1})$.
:::

::: {#prp-ffrcistg-markov}
## Counterfactual Models Imply the Causal Markov Assumption

If the counterfactuals for a causal DAG $G$ follow an FFRCISTG (@def-ffrcistg),
and so in particular if they follow an NPSEM-IE (@prp-npsem-ie-ffrcistg),
then the joint distribution of the factual variables is Markov with respect to $G$ (@def-markov-dag).
Robins (1986) proved this [@hernan2020causal, Technical Point 6.3, p. 78].
:::

::: {.callout-note title="The Default Counterfactual Model"}
Unless stated otherwise, a causal DAG in the book, and in these notes, represents an FFRCISTG (@def-ffrcistg)
[@hernan2020causal, Technical Point 6.3, p. 78].
:::

::: {.notes}
Robins and Richardson (2010) showed that Robins's original formulation is equivalent to the joint-independence assumption for a positive distribution,
and that an NPSEM-IE makes many more independence assumptions than an FFRCISTG.
:::

## 6.2 Causal Diagrams and Marginal Independence (pp. 77-79)

---

Three examples, all with no conditioning:

```{r}
#| label: fig-dag-6-3
#| echo: false
#| fig-width: 4
#| fig-height: 2.5
#| fig-cap: "Book Figure 6.3: smoking $L$ causes carrying a lighter $A$ and lung cancer $Y$."
draw_dag('dag {
  L [pos="1,1"]
  A [pos="0,0"]
  Y [pos="2,0"]
  L -> A
  L -> Y
}')
```

```{r}
#| label: fig-dag-6-4
#| echo: false
#| fig-width: 4
#| fig-height: 2.5
#| fig-cap: "Book Figure 6.4: haplotype $A$ and smoking $Y$ both cause heart disease $L$, a collider."
draw_dag('dag {
  A [pos="0,0"]
  L [pos="1,0"]
  Y [pos="2,0"]
  A -> L
  Y -> L
}')
```

::: {#tbl-marginal-examples}

| Figure | Structure | Example | $A$ and $Y$ |
|---|---|---|---|
| @fig-dag-6-2 | $A \to Y$ | aspirin prevents heart disease (randomized) | associated |
| @fig-dag-6-3 | $A \leftarrow L \to Y$ | lighter, smoking, lung cancer | associated (no effect of $A$) |
| @fig-dag-6-4 | $A \to L \leftarrow Y$ | haplotype, heart disease, smoking | independent |

Marginal association between $A$ and $Y$ in three causal DAGs.
"Associated" holds generically (for all but special parameter values).
:::

---

### Paths and the Flow of Association

::: {#def-path}
## Path; Causal Path

A **path** between $R$ and $S$ is a route connecting them along a sequence of edges that visits no variable more than once.
A path is **causal** if all its arrows point in the same direction;
otherwise it is **noncausal**
[@hernan2020causal, p. 78, margin note].
:::

::: {#exm-paths-6-1}
## Paths in Book Figure 6.1

In @fig-dag-6-1 there are two paths between $A$ and $Y$:
$A \to Y$, which is causal,
and $A \leftarrow L \to Y$, which is noncausal because its two arrows point in different directions.
:::

Picture paths as pipes through which association flows.
Association is symmetric, so it flows regardless of the direction of the arrows.

---

::: {#def-collider}
## Collider

A variable on a path is a **collider** on that path if two arrowheads on the path collide at it,
as $L$ does on $A \to L \leftarrow Y$.
:::

::: {#exm-collider-6-1}
## Being a Collider Depends on the Path

In @fig-dag-6-1, $Y$ is a collider on the path $A \to Y \leftarrow L$ between $A$ and $L$.
On the path $L \to A \to Y$ between $L$ and $Y$, by contrast,
$A$ is not a collider: one arrow on the path enters it and the other leaves it.
:::

---

### The Three Examples

::: {#exm-three-marginal-examples}
## Three Marginal Associations

- @fig-dag-6-2: causation implies association; in an ideal randomized experiment
  $\Pr[Y^{a=1}=1] \neq \Pr[Y^{a=0}=1]$ if and only if $\Pr[Y=1 \mid A=1] \neq \Pr[Y=1 \mid A=0]$.
- @fig-dag-6-3: learning that Hera carries a lighter makes it likelier she smokes, hence likelier she gets lung cancer;
  association flows through the common cause $L$, though $A$ has no effect on $Y$.
- @fig-dag-6-4: learning that Apollo lacks the haplotype says nothing about his smoking;
  the collider $L$ blocks the flow of association, so $A \ind Y$.
:::

::: {#rem-marginal-association}
## Marginal Association in a Causal DAG

Two variables on a causal DAG are generically (for all but special parameter values) marginally associated
if one causes the other or if they share a common cause.
Otherwise, under the causal Markov assumption (@def-causal-markov), they are marginally independent.
:::

::: {.notes}
In @fig-dag-6-3, an investigator who concludes from the association that carrying a lighter causes lung cancer makes a mistake:
information about $A$ improves prediction of $Y$ without $A$ affecting $Y$ [@hernan2020causal, p. 79].
In @fig-dag-6-4, that both $A$ and $Y$ cause heart disease $L$ is irrelevant to the marginal association between $A$ and $Y$.
:::

## 6.3 Causal Diagrams and Conditional Independence (pp. 80-81)

---

::: {.callout-note title="Notation: A Box Means Conditioning"}
A square box around a node means that the analysis conditions on that variable,
for example by restricting to one of its levels.
:::

```{r}
#| label: fig-dag-6-5
#| echo: false
#| fig-width: 4
#| fig-height: 1.5
#| fig-cap: "Book Figure 6.5: aspirin $A$ lowers platelet aggregation $B$, which affects heart disease $Y$ (the square node $B$ marks conditioning)."
draw_dag('dag {
  A [pos="0,0"]
  B [pos="1,0"]
  Y [pos="2,0"]
  A -> B
  B -> Y
}', conditioned = "B")
```

::: {#exm-conditioning-three-structures}
## Conditioning in Three Structures

| Structure | Conditioning on | Path | $A$, $Y$ given the conditioning variable |
|---|---|---|---|
| $A \to B \to Y$ (Fig. 6.5) | mediator $B$ | blocked | independent: $A \ind Y \mid B$ |
| $A \leftarrow L \to Y$ (Fig. 6.6) | common cause $L$ | blocked | independent: $A \ind Y \mid L$ |
| $A \to L \leftarrow Y$ (Fig. 6.7) | collider $L$ | **opened** | generically associated |

The two independences follow from the causal Markov assumption (@def-causal-markov);
"generically" means for all but special parameter values.
:::

::: {.notes}
- **Mediator** [@hernan2020causal, p. 80]: among those with low platelet aggregation ($B=0$),
  knowing whether someone took aspirin adds no information about heart disease,
  because aspirin acts only through $B$:
  $\Pr[Y=1 \mid A=1, B=b] = \Pr[Y=1 \mid A=0, B=b]$ for all $b$.
- **Common cause**: among nonsmokers ($L=0$), carrying a lighter no longer predicts lung cancer.
  Blocking the flow of association through common causes is the graph-based justification
  for stratification as a way to achieve exchangeability.
- Because complete diagrams (all possible arrows present) imply no conditional independences,
  it is often said that the information about associations is in the **missing** arrows.
:::

---

### Conditioning on a Collider Opens the Path

::: {#exm-collider-heart-disease}
## Haplotype and Smoking Among People with Heart Disease

In @fig-dag-6-4, restrict to people with heart disease ($L=1$, book Figure 6.7).
Intuitively, among them someone without the haplotype is more likely to have the other cause, smoking,
which would make $A$ and $Y$ inversely associated given $L=1$.
In the extreme case where $L = 1$ exactly when $A = 1$ or $Y = 1$,
among those with $L=1$ the absence of one cause guarantees the presence of the other.
:::

::: {.callout-warning title="The Diagram Does Not Fix the Sign"}
The diagram alone implies only that, generically (for all but special parameter values),
$A$ and $Y$ are associated in at least one level of $L$.
The sign of that association, and whether it appears in the level $L = 1$ in particular,
depend on how $A$ and $Y$ act together on $L$ (Chapter 8).
:::

::: {.notes}
Intuition from the book [@hernan2020causal, p. 81]:
whether two causes are associated cannot depend on a future event (their common effect),
but two causes of an effect generally become associated once we stratify on that effect.
Chapter 8 develops associations due to conditioning on common effects.
:::

---

### Conditioning on a Descendant of a Collider

```{r}
#| label: fig-dag-6-8
#| echo: false
#| fig-width: 4
#| fig-height: 2.5
#| fig-cap: "Book Figure 6.8: diuretic use $C$ follows a diagnosis of heart disease $L$ (the square node $C$ marks conditioning)."
draw_dag('dag {
  A [pos="0,0"]
  L [pos="1,0"]
  Y [pos="2,0"]
  C [pos="1,1"]
  A -> L
  Y -> L
  L -> C
}', conditioned = "C")
```

::: {#exm-descendant-of-collider}
## Conditioning on Diuretic Use

In @fig-dag-6-8, diuretic use $C$ is affected by the collider heart disease $L$.
Conditioning on $C$ also opens the path $A \to L \leftarrow Y$,
so haplotype and smoking are generically (for all but special parameter values) associated within levels of $C$.
The path stays blocked when we condition on neither $L$ nor $C$.
:::

---

### Three Structural Sources of Association

::: {#rem-structural-sources-association}
## Three Structural Sources of Association

Two variables can be associated because

1. one causes the other;
2. they share a common cause;
3. they share a common effect and the analysis is restricted to a level of that effect (or of its descendants).

A fourth, nonstructural source is chance (random variability), which shrinks as the study population grows.
Until Chapter 10 the book assumes a very large population, so all associations discussed are structural.
:::

::: {.notes}
The heuristic arguments of this section are formalized by d-separation (Pearl 1995);
Fine Point 6.1 lists the rules and Fine Point 6.2 introduces faithfulness [@hernan2020causal, p. 81].
:::

## 6.4 Positivity and Consistency in Causal Diagrams (pp. 81-85)

---

::: {.callout-note title="Fine Point 6.1: D-Separation"}
The graphical rules of Sections 6.2 and 6.3, for a path (@def-path) between two variables, are:

1. with nothing conditioned on, a path is blocked exactly when it contains a collider (@def-collider);
2. a path that contains a conditioned-on non-collider is blocked;
3. a conditioned-on collider does not block a path;
4. a collider with a conditioned-on descendant does not block a path.

The two definitions below summarize these rules [@hernan2020causal, Fine Point 6.1, p. 82].
:::

::: {#def-blocked-path}
## Blocked and Open Paths

Given a set $\mathcal{C}$ of conditioned-on variables, a path is **blocked** if and only if

- it contains a non-collider that is in $\mathcal{C}$, or
- it contains a collider that is not in $\mathcal{C}$ and has no descendant in $\mathcal{C}$.

Otherwise the path is **open**.
:::

::: {#def-d-separation}
## D-Separation

Two variables are **d-separated** given a set $\mathcal{C}$ of other variables ($\mathcal{C}$ may be empty)
if every path between them is blocked given $\mathcal{C}$ (@def-blocked-path);
otherwise they are **d-connected** given $\mathcal{C}$.
Two disjoint sets of variables are d-separated given $\mathcal{C}$ if each variable in one is d-separated from every variable in the other.
:::

::: {#exm-d-separation}
## D-Separation in Book Figures 6.1, 6.4, and 6.8

- @fig-dag-6-1, $A$ and $L$, given nothing: the path $L \to A$ is open, so they are d-connected,
  even though the other path $A \to Y \leftarrow L$ is blocked by the collider $Y$.
- @fig-dag-6-4, $A$ and $Y$, given nothing: the only path is blocked by the collider $L$, so they are d-separated.
- @fig-dag-6-8, $A$ and $Y$, given $\{C\}$: the collider $L$ has the descendant $C$ in the conditioning set,
  so the only path is open and they are d-connected.
:::

---

### From D-Separation to Independence

::: {#thm-d-separation}
## D-Separation Implies Independence

Let $\mathcal{A}$, $\mathcal{B}$, $\mathcal{C}$ be disjoint sets of variables on a causal DAG,
and suppose the causal Markov assumption (@def-causal-markov) holds.
If $\mathcal{A}$ and $\mathcal{B}$ are d-separated given $\mathcal{C}$ (@def-d-separation),
then $\mathcal{A} \ind \mathcal{B} \mid \mathcal{C}$.
Pearl (1988) proved this [@hernan2020causal, Fine Point 6.1, p. 82].
:::

::: {#exm-d-separation-independence}
## Reading an Independence off Book Figure 6.5

In @fig-dag-6-5, the only path between $A$ and $Y$ is $A \to B \to Y$,
and it contains the conditioned-on non-collider $B$.
So $A$ and $Y$ are d-separated given $\{B\}$,
and @thm-d-separation gives $A \ind Y \mid B$,
the conclusion reached informally in Section 6.3.
:::

---

::: {#def-faithfulness}
## Faithfulness

The joint distribution of the variables on a causal DAG is **faithful** to the DAG if,
for all disjoint sets $\mathcal{A}$, $\mathcal{B}$, $\mathcal{C}$ of its variables ($\mathcal{C}$ possibly empty),
$\mathcal{A} \ind \mathcal{B} \mid \mathcal{C}$ implies that $\mathcal{A}$ and $\mathcal{B}$ are d-separated given $\mathcal{C}$:
the converse of @thm-d-separation.
The **faithfulness assumption** is that this holds
[@hernan2020causal, Fine Point 6.2, p. 83].
:::

::: {#exm-faithfulness-collider}
## Faithfulness and a Conditioned-On Collider

In @fig-dag-6-4 with $L$ conditioned on (book Figure 6.7), $A$ and $Y$ are d-connected given $\{L\}$.
Under faithfulness, then, $A$ and $Y$ are not independent given $L$:
they are associated in at least one level of $L$.
:::

::: {.notes}
"d-" stands for directional.
An equivalent set of graphical rules, **moralization**, was developed by Lauritzen et al. (1990)
[@hernan2020causal, Fine Point 6.1, p. 82].
:::

---

::: {.callout-note title="Fine Point 6.2: Faithfulness"}
An arrow $A \to Y$ means $A$ affects $Y$ for at least one individual,
yet the average causal effect and the association can still both be null,
when effects in different individuals cancel exactly.
Such exact cancellation is rare, so the book assumes faithfulness (@def-faithfulness):
lack of d-separation can almost always be equated with a non-zero association.
:::

::: {#exm-unfaithful-table-4-1}
## Unfaithfulness from Cancelling Effects

In the book's Table 4.1, transplant increases the risk of death in women and decreases it in men,
each half of the population, and the effects cancel exactly:
$\Pr[Y^{a=1}=1] = \Pr[Y^{a=0}=1]$.
@fig-dag-6-2 is still the correct diagram, because $A$ affects every individual's $Y$,
but the expected association is absent: the distribution is not faithful to the DAG (@def-faithfulness)
[@hernan2020causal, Fine Point 6.2, p. 83].
:::

::: {#exm-matching-unfaithful}
## Matching Builds In a Cancellation

In a matched study (book Section 4.5, Figure 6.9), the matching factor $L$ affects treatment $A$,
selection $S$ depends on both $A$ and $L$,
and the analysis is restricted to the matched individuals ($S=1$).
Given $S$, the paths $L \to A$ and $L \to S \leftarrow A$ are both open, so $L$ and $A$ are d-connected given $\{S\}$.
Yet matching makes the distribution of $L$ the same in the treated and the untreated,
so $L$ and $A$ are independent among the matched:
the association created through $L \to S \leftarrow A$ exactly offsets the one through $L \to A$.
This is the kind of exact cancellation that faithfulness rules out,
which is why the book says that matching leads to unfaithfulness [@hernan2020causal, Fine Point 6.2, p. 83].
:::

::: {.notes}
Faithfulness may also fail when variables are linked by deterministic arrows:
two variables can then be independent even though some paths between them are open.
:::

---

### Graphs, Do-Calculus, and Notation

Standardization and IP weighting (Chapter 2) can also be derived from causal graph theory,
as part of what is sometimes called the do-calculus (reviewed in Pearl 2009).
So using counterfactuals in Chapters 1-5 privileged a notation, not an approach.

Whatever the notation, causal inference via standardization or IP weighting requires
exchangeability, positivity, and consistency.
This section covers positivity and consistency;
exchangeability is translated into graphs in Section 6.5 and Chapters 7-8.

---

### Positivity and Consistency in the Graph

::: {#rem-positivity-consistency-graph}
## Where Positivity and Consistency Sit in a Graph

- **Positivity** concerns arrows **into** treatment nodes.
  Causal graphs generally cannot encode its violations, except special cases,
  e.g., $A$ a deterministic function of a pretreatment $L$ (drawn as a bold $L \to A$ arrow).
- **Consistency** (well-defined counterfactuals) concerns arrows **out of** treatment nodes:
  $A \to Y$ must correspond to a possibly hypothetical but relatively unambiguous intervention
  (or, if $A$ is a state, to well-defined interventions that affect $Y$ only through $A$;
  Fine Point 6.4).

In the book's diagrams, positivity is implicit unless stated
and consistency is embedded in the notation.
:::

::: {.notes}
Treatment nodes thus have a special status [@hernan2020causal, pp. 83-84].
Some authors draw it explicitly with **decision nodes**;
influence diagrams are causal diagrams augmented with decision nodes (Dawid 2000, 2002).
The book omits them, because it is always explicit about the interventions on $A$,
but gives treatment nodes a distinct status in SWIGs later.
The causal trees of Chapter 2 and the sufficient-cause "pies" of Chapter 5 also distinguished treatments from other variables.
:::

---

### Treatment Nodes Need Well-Defined Interventions

::: {.callout-warning title="Not Every Node Can Be a Treatment"}
Diagrams read as NPSEMs with independent errors (@def-npsem-ie) seem to give all variables equal status,
which can mislead when some nodes correspond to ill-defined interventions.

- A node for "obesity" may be acceptable as an outcome $Y$ or a covariate $L$.
- It is generally **not** acceptable as a treatment $A$ (Chapter 3).
:::

::: {.notes}
Pearl (2018, 2019) has proposed a concept of causation based on variables that "listen to others",
which still assumes well-defined counterfactuals for every variable [@hernan2020causal, p. 84, margin note].
:::

---

```{r}
#| label: fig-dag-6-10
#| echo: false
#| fig-width: 5
#| fig-height: 3
#| fig-cap: "Book Figure 6.10, as described in the text: caloric intake $Z$, exercise $L$, and genetic traits $U$ affect weight loss $A$; $L$ and $U$ also affect mortality $Y$ through other pathways."
draw_dag('dag {
  Z [pos="0,0"]
  L [pos="0,1"]
  U [pos="0,-1"]
  A [pos="1,0"]
  Y [pos="2,0"]
  Z -> A
  L -> A
  U -> A
  A -> Y
  L -> Y
  U -> Y
}')
```

::: {#exm-weight-loss-interventions}
## Many Ways to Lose Weight

In @fig-dag-6-10 there are several ways to intervene on weight loss $A$,
and some of them, through exercise $L$ or genetic traits $U$, also affect mortality $Y$ directly.
Losing weight by caloric restriction, by exercise, or by genetic manipulation would then lead to different mortality,
so "the effect of $A$ on $Y$" is unclear.
:::

::: {.callout-tip title="Name the Intervention"}
Being explicit about the intervention is a step toward a well-defined effect,
relevant data, and the right adjustment variables.
:::

---

::: {#def-causal-discovery}
## Causal Discovery

**Causal discovery** is learning parts of the causal structure, such as which arrows are absent from the causal DAG,
by analyzing data rather than from expert knowledge [@hernan2020causal, Fine Point 6.3, p. 85].
:::

::: {#exm-discovery}
## Learning $Z \to A \to Y$

Suppose $Z$ precedes $A$, which precedes $Y$,
and assume faithfulness (@def-faithfulness).
With infinite data, suppose all three variables are marginally associated
and the only conditional independence is $Z \ind Y \mid A$.
Then the only causal DAGs compatible with the data are $Z \to A \to Y$,
possibly with a common cause of $Z$ and $A$ in addition to, or instead of, $Z \to A$.
So there is no unmeasured common cause of $A$ and $Y$,
and, if $Y^a$ is well defined and positivity holds,
the average causal effect is identified by $\E{Y \mid A=1} - \E{Y \mid A=0}$
[@hernan2020causal, Fine Point 6.3, p. 85].
:::

::: {.callout-note title="Fine Point 6.3: Discovery of Causal Structure"}
Discovery (@def-causal-discovery) sometimes works if we assume faithfulness,
so that independences in the data imply missing arrows,
but often it does not:
a strong association between $B$ and $C$ fits $B \to C$, $C \to B$, a shared unmeasured cause,
a conditioned-on common effect, and combinations of these.
:::

::: {.notes}
Why the learned DAG in @exm-discovery: a direct arrow $Z \to Y$, a common cause of $Z$ and $Y$, or an unmeasured common cause of $A$ and $Y$
would each make $Z$ and $Y$ dependent given $A$ (assuming faithfulness);
the marginal association of $A$ and $Y$ then requires an arrow $A \to Y$.
If an unmeasured common cause of $A$ and $Y$ existed, no conditional independence would be found,
and discovery could not tell whether $A$ causes $Y$.
Approaches to discovery are described by Spirtes et al. (2000) and Peters et al. (2017);
finite samples are discussed in the book's Technical Point 10.7.
:::

## 6.5 A Structural Classification of Bias (pp. 85-87)

---

::: {#def-systematic-bias}
## Systematic Bias

There is **systematic bias** when the data are insufficient to identify the causal effect even with an infinite sample size.
Informally: any structural association between treatment and outcome that does not arise from the causal effect of treatment on outcome
[@hernan2020causal, p. 85].
:::

::: {#exm-systematic-bias-lighter}
## Lighters and Lung Cancer Again

In @fig-dag-6-3, carrying a lighter $A$ has no effect on lung cancer $Y$,
yet the two are generically (for all but special parameter values) associated through smoking $L$.
That association is structural: it does not shrink as the sample grows.
So even infinite data on $A$ and $Y$ alone cannot reveal that the effect is null:
an analysis that ignores $L$ has systematic bias.
:::

::: {.notes}
With systematic bias no estimator can be consistent (see Chapter 1 for consistent estimators).
In this chapter "bias" means systematic bias, because the sample size is assumed infinite.
:::

---

::: {#def-unconditional-conditional-bias}
## Unconditional and Conditional Bias

For a dichotomous treatment $A$, a dichotomous outcome $Y$, and a covariate $L$:

- there is **unconditional bias** when
  $\Pr[Y^{a=1}=1] - \Pr[Y^{a=0}=1] \neq \Pr[Y=1 \mid A=1] - \Pr[Y=1 \mid A=0]$;
- there is **conditional bias** given $L$ when
  $\Pr[Y^{a=1}=1 \mid L=l] - \Pr[Y^{a=0}=1 \mid L=l] \neq \Pr[Y=1 \mid L=l, A=1] - \Pr[Y=1 \mid L=l, A=0]$
  for at least one $l$.
:::

Under consistency and positivity, unconditional exchangeability $Y^a \ind A$ rules out unconditional bias,
and conditional exchangeability $Y^a \ind A \mid L$ rules out conditional bias.
When exchangeability fails, bias is generically present.
Absence of unconditional bias means the association measure in the population is a consistent estimate of the corresponding effect measure.

::: {#exm-table-3-1-bias}
## Unconditional but No Conditional Bias in Table 3.1

In the book's Table 3.1, assume, as in Chapter 3, consistency, positivity, and conditional exchangeability given $L$.
Within each level of $L$ the observed risks are equal in the treated and the untreated
(1/4 versus 1/4 when $L=0$, 2/3 versus 2/3 when $L=1$),
so the conditional causal risk differences are 0 and there is no conditional bias.
The causal risk difference is also 0,
but the associational risk difference is $7/13 - 3/7 \approx 0.11$:
there is unconditional bias.
:::

---

### Bias Under the Null

::: {#def-bias-under-null}
## Bias Under the Null

There is **bias under the null** when treatment has no causal effect on the outcome
and yet, in the population, treatment and outcome are associated.
Lack of exchangeability causes bias under the null.
:::

::: {#exm-bias-under-null}
## Bias Under the Null in Table 3.1

In the book's observational study of Table 3.1 (@exm-table-3-1-bias), the causal risk ratio was 1,
whereas the associational risk ratio was 1.26 [@hernan2020causal, p. 86].
:::

::: {#rem-bias-null-alternative}
## Under the Null and Under the Alternative

Any causal structure that causes bias under the null also causes bias under the alternative,
that is, when treatment does have a non-null effect;
the converse is false.
:::

::: {.notes}
For example, conditioning on some variables may cause selection bias under the alternative but not under the null
(Greenland 1977; Hernán 2017;
see also Chapter 18) [@hernan2020causal, p. 86, margin note].
:::

---

### Two Structures That Break Exchangeability

::: {#rem-sources-systematic-bias}
## Sources of Systematic Bias

Lack of exchangeability has two causal structures:

1. common causes of treatment and outcome: what many epidemiologists call confounding (Chapter 7);
2. conditioning on common effects: what many epidemiologists call selection bias under the null (Chapter 8).

A third source of systematic bias is measurement bias (information bias), from mismeasured treatment, outcome, or covariates;
some types also cause bias under the null (Chapter 9).
Random variability is a separate, nonstructural source of error (Chapter 10).
:::

::: {.notes}
All three biases can arise in observational studies **and** in randomized experiments.
Earlier chapters used idealized experiments (no loss to follow-up, full adherence, blinded assignment),
but real experiments rarely look like that;
the remaining chapters of Part I examine the boundary between experimenting and observing [@hernan2020causal, pp. 86-87].
:::

## 6.6 The Structure of Effect Modification (pp. 87-89)

---

Causal diagrams are good at locating sources of bias,
but less helpful for illustrating effect modification.

```{r}
#| label: fig-dag-6-12
#| echo: false
#| fig-width: 4
#| fig-height: 2.5
#| fig-cap: "Book Figure 6.12: randomized transplant $A$; quality of care $V$ affects death $Y$."
draw_dag('dag {
  A [pos="0,0"]
  V [pos="1,1"]
  Y [pos="2,0"]
  A -> Y
  V -> Y
}')
```

::: {#exm-quality-of-care}
## Transplant and Quality of Care

In @fig-dag-6-12, transplant $A$ is randomized, so association is causation.
Investigators stratify by quality of care $V$ and find effect modification by $V$ on the additive scale.
:::

---

### Two Caveats

::: {.callout-warning title="What the Diagram Does Not Say About Effect Modification"}
1. @fig-dag-6-12 would be valid **without** $V$, since $V$ is not a common cause of $A$ and $Y$;
   $V$ appears only because the question refers to it.
   Variables on the path from $V$ to $Y$, such as therapy complications $N$ (book Figure 6.13), could also be effect modifiers.
2. @fig-dag-6-12 does **not** show whether, or how, $V$ modifies the effect of $A$.
   It cannot distinguish:
   - effects in the same direction in both strata of $V$;
   - effects in opposite directions (qualitative effect modification);
   - an effect in one stratum only (e.g., $A$ kills only those with $V=0$).
:::

---

### Surrogate Effect Modifiers

::: {#def-surrogate-effect-modifier}
## Causal and Surrogate Effect Modifiers

A **causal effect modifier** is a variable that modifies the effect of treatment on the outcome
([Chapter 4](04-effect-modification.qmd)) and itself affects the outcome.
A **surrogate effect modifier** is a variable that is associated with a causal effect modifier but does not itself affect the outcome
[@hernan2020causal, p. 88].
:::

::: {#exm-surrogate-modifiers}
## Three Surrogates for Quality of Care

In the study of @exm-quality-of-care, an analysis stratified by a surrogate of quality of care $V$, but not by $V$ itself,
will generally detect effect modification by the surrogate.
The association between surrogate and $V$ can come from any structure:

| Book figure | Surrogate | Structure linking it to $V$ |
|---|---|---|
| 6.14 | cost of treatment $S$ | cause and effect: $V \to S$ |
| 6.15 | passport nationality $P$ | common cause: residence $U$ affects $V$ and $P$ |
| 6.16 | bottled mineral water $W$ | conditioning on a common effect: $V$ and $W$ affect cost $S$, analysis restricted to $S=0$ |
:::

::: {.callout-warning title="A Surrogate Is Not a Lever"}
Reading surrogate effect modification causally misleads.
"Cost modifies the effect of transplant" could suggest raising prices without raising the quality of care
[@hernan2020causal, p. 88].
:::

::: {.notes}
Causal and surrogate effect modifiers are often indistinguishable in practice,
so "effect modification" covers both (Section 4.2);
some prefer the neutral term "heterogeneity of causal effects".

Intuition for the mineral-water example (book margin note, p. 88):
low-cost hospitals that buy mineral water must spend less on care that lowers mortality,
so among low-cost hospitals, mineral water use is inversely associated with quality of care.
VanderWeele and Robins (2007b) give a finer classification of effect modification via causal diagrams.
:::

---

### Interaction in Causal Diagrams

::: {#rem-diagrams-interaction}
## Diagrams and Interaction

Causal diagrams are agnostic about interaction between two treatments $A$ and $E$.
They can encode it if augmented with nodes for sufficient-component causes (Chapter 5),
with deterministic arrows from the treatments to those nodes;
the book develops such diagrams in Chapter 8.
:::

---

::: {.callout-note title="Fine Point 6.4: Evidence That a State Has Well-Defined Counterfactuals"}
Consider drug $Z$, systolic blood pressure $A$ (a state, not directly manipulable), and stroke $Y$.
If (i) $Z$ is associated with $A$ and $Y$, (ii) $A$ and $Y$ are associated, and (iii) $Z \ind Y \mid A$,
then under faithfulness the only causal DAG has $A \to Y$, no $Z \to Y$, and no unmeasured common cause of $A$ and $Y$ or of $Z$ and $Y$
(the discovery argument of @exm-discovery).

- If $Y^a$ is well defined and positivity holds, $Y^a \ind A$ holds and $\E{Y \mid A=1} - \E{Y \mid A=0}$ is the average causal effect.
- An investigator who doubted that $Y^a$ was well defined might take (i)-(iii) as empirical evidence that it is.

::: {.notes}
Usually unmeasured common causes $U_3$ of $A$ and $Y$ remain;
then $Z$ and $Y$ are associated given the collider $A$, (iii) fails,
and the data give no support for $Y^a$ being well defined [@hernan2020causal, Fine Point 6.4, p. 89].

The book's remedy is a four-arm factorial **dose-finding** trial:
randomize the drug ($Z=1$ or $Z=2$) and the target blood-pressure reduction ($A=a_1$ or $A=a_2$),
and increase each individual's dose $D$ until the assigned reduction is reached and maintained.
The achieved $A$ is then a deterministic function of the assigned value, so nothing else has an arrow into $A$ (book Figure 6.18).
If $Y$ is independent of $D$ given $A$ and $\E{Y \mid A=a_1} \neq \E{Y \mid A=a_2}$,
then under faithfulness (and assuming $Y^{d,a}$ and $Y^a$ are well defined)
$D$ has no direct effect on $Y$ outside $A$,
and $\E{Y^{a_1} - Y^{a_2}}$ is identified by $\E{Y \mid A=a_1} - \E{Y \mid A=a_2}$,
even with unknown common causes of $A$ and $Y$.
In a standard trial (book Figure 6.17), by contrast, $Y$ and $Z$ are associated given $A$,
so a direct effect of $Z$ on $Y$ cannot be ruled out.
:::
:::

## Summary

---

- A **causal DAG** encodes qualitative causal knowledge:
  an arrow means a direct effect for at least one individual;
  a missing arrow means none for anyone;
  all common causes of variables on the graph must be on the graph (causal Markov assumption).
- Each causal DAG represents a counterfactual model (by default an FFRCISTG).
- Association flows along open paths:
  marginally, causal chains and common causes transmit association, colliders block it;
  conditioning on a non-collider blocks a path, conditioning on a collider or its descendant opens it (**d-separation**).
- **Faithfulness**, assumed throughout the book, lets us equate d-connection with association.
- Graphs rarely show **positivity** violations;
  **consistency** requires treatment nodes with well-defined interventions.
- **Systematic bias** from lack of exchangeability has two structures: common causes (confounding) and conditioning on common effects (selection bias);
  measurement bias is a third type.
- Causal diagrams do not show whether or how a variable modifies an effect, and cannot separate causal from surrogate effect modifiers.

::: {.notes}
**Looking ahead**: Chapter 7 (confounding), Chapter 8 (selection bias), Chapter 9 (measurement bias), Chapter 10 (random variability).

The diagrams in this chapter are drawn with the R packages `dagitty` and `ggdag`.
:::

## References

---

::: {#refs}
:::
