---
title: "Chapter 7: Confounding"
format:
  html: default
  revealjs:
    output-file: 07-confounding-slides.html
  pdf:
    output-file: 07-confounding-handout.pdf
  docx:
    output-file: 07-confounding.docx
---

{{< include ../latex-macros/macros.qmd >}}

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
library(ggdag)
library(dagitty)
library(ggplot2)
library(patchwork)

# Draw a causal DAG; nodes named in `unmeasured` are shown in gray.
draw_dag <- function(dag, unmeasured = character(), title = NULL) {
  ggdag(dag, node = FALSE, text = FALSE, stylized = TRUE) +
    theme_dag() +
    geom_dag_point(aes(color = name %in% unmeasured), size = 14) +
    scale_color_manual(
      values = c("FALSE" = "steelblue", "TRUE" = "gray60"),
      guide = "none"
    ) +
    geom_dag_text(color = "white", size = 4) +
    labs(title = title)
}
```

::: {#exm-looking-up-thunder}
## Looking Up at the Sky, with Thunder

Suppose an investigator observes that when one pedestrian looks up at the sky, a second pedestrian tends to look up too.
She also notices that pedestrians look up when they hear a thunderous noise overhead,
so she cannot tell whether the second pedestrian looked up because of the first one or because of the noise:
the noise drives both looks,
so the effect of one person's looking up is entangled with the effect of the noise.
:::

In a randomized experiment treatment is assigned by a coin flip,
but in an observational study treatment may be determined by factors that also affect the outcome.
The effects of those factors then become entangled with the effect of treatment.
This chapter calls that entanglement *confounding* and defines it structurally,
links it to exchangeability,
contrasts the structural definition with the traditional definition of a confounder,
introduces single-world intervention graphs (SWIGs),
and reviews methods to adjust for confounding.

::: {.notes}
This chapter is based on @hernan2020causal [Chapter 7, pp. 91-107].

The book describes confounding as "just a form of lack of exchangeability between the treated and the untreated" [@hernan2020causal, p. 91].
In the presence of confounding, "association is not causation" holds no matter how large the study population is.
:::

## 7.1 The Structure of Confounding (pp. 91-93)

---

```{r}
#| label: dag-ch7-1
#| echo: false
#| fig-width: 4
#| fig-height: 2.5
#| fig-cap: "Figure 7.1 (same as Figure 6.1). L is a common cause of treatment A and outcome Y."
dag_7_1 <- dagitty('dag {
  L [pos="0,1"]; A [pos="1,0"]; Y [pos="2,0"]
  L -> A; L -> Y; A -> Y
}')
draw_dag(dag_7_1)
```

In Figure 7.1 there are two sources of association between $A$ and $Y$:

- the causal path $A \rightarrow Y$;
- the path $A \leftarrow L \rightarrow Y$ through the common cause $L$.

The second path enters $A$ through an arrow pointing into $A$.
Such paths have a name.
(Paths, causal paths and colliders are defined in [Chapter 6](06-graphical-representation.qmd#def-path).)

::: {#def-backdoor-path-ch7}
## Backdoor Path

In a causal DAG, a **backdoor path** is a noncausal path between treatment and outcome that remains even if all arrows pointing from treatment to other variables (the descendants of treatment) are removed.
That is, the path has an arrow pointing into treatment [@hernan2020causal, p. 91].
:::

::: {#exm-backdoor-path-7-1}
## The Backdoor Path in Figure 7.1

Deleting the arrow $A \rightarrow Y$ from Figure 7.1 leaves $A \leftarrow L \rightarrow Y$ in place,
and that path starts with an arrow into $A$,
so it is a backdoor path.
The path $A \rightarrow Y$ is causal, so it is not a backdoor path.
:::

---

If $L$ did not exist, all the association between $A$ and $Y$ would be causal,
and the associational risk ratio $\Pr[Y = 1 \mid A = 1] / \Pr[Y = 1 \mid A = 0]$ would equal the causal risk ratio $\Pr[Y^{a=1} = 1] / \Pr[Y^{a=0} = 1]$.
The common cause $L$ adds a second source of association.

::: {#def-confounding}
## Confounding

Let $A$ be a treatment and $Y$ an outcome in a causal DAG.
**Confounding** for the effect of $A$ on $Y$ is the bias that results from common causes of $A$ and $Y$,
which open backdoor paths (@def-backdoor-path-ch7) between them.
Here *bias* means [systematic bias](06-graphical-representation.qmd#def-systematic-bias):
the association between $A$ and $Y$ differs from their causal effect even in an infinitely large population.
There is **no confounding** when $A$ and $Y$ share no common cause, measured or unmeasured.
:::

::: {#exm-confounding-7-1}
## Confounding in Figure 7.1

In Figure 7.1, $L$ is a common cause of $A$ and $Y$.
Suppose that, for the sake of illustration, $L$ raises both the chance of treatment and the risk of the outcome, and $A$ has no effect on $Y$.
Then the treated have more individuals with $L = 1$ than the untreated, and so a higher risk,
and the associational risk ratio exceeds the causal risk ratio of 1.
The gap between them is confounding.
:::

### Examples of Confounding

---

```{r}
#| label: dag-ch7-2-3
#| echo: false
#| fig-width: 8
#| fig-height: 3
#| fig-cap: "Left: Figure 7.2, in which L causes A, and L and Y share the unmeasured cause U. Right: Figure 7.3, in which L causes Y, and L and A share the unmeasured cause U. Unmeasured nodes are gray."
dag_7_2 <- dagitty('dag {
  U [pos="1,1"]; L [pos="0,1"]; A [pos="1,0"]; Y [pos="2,0"]
  U -> L; U -> Y; L -> A; A -> Y
}')
dag_7_3 <- dagitty('dag {
  U [pos="0,0.5"]; L [pos="1,1"]; A [pos="1,0"]; Y [pos="2,0"]
  U -> A; U -> L; L -> Y; A -> Y
}')
draw_dag(dag_7_2, unmeasured = "U", title = "Figure 7.2") +
  draw_dag(dag_7_3, unmeasured = "U", title = "Figure 7.3")
```

::: {#exm-confounding-in-practice}
## Confounding Across Fields

- **Occupational factors** (Figure 7.1):
  physical fitness $L$ causes both working as a firefighter $A$ and lower mortality $Y$ (**healthy worker bias**).
- **Clinical decisions** (Figure 7.1 or 7.2):
  heart disease $L$ is an indication for aspirin $A$ and a risk factor for stroke $Y$,
  either directly or through unmeasured atherosclerosis $U$ (**confounding by indication** or **channeling**).
- **Lifestyle** (Figure 7.3):
  personality and social factors $U$ lead to both lack of exercise $A$ and smoking $L$, which affects death $Y$.
- **Genetic factors** (Figure 7.3):
  a DNA sequence $L$ that affects trait $Y$ is more frequent among carriers of sequence $A$
  (**linkage disequilibrium** or **population stratification**).
- **Social factors** (Figure 7.1):
  disability at age 55 $L$ affects income at age 65 $A$ and disability at age 75 $Y$.
- **Environmental exposures** (Figure 7.3):
  weather $U$ affects levels of particulate matter $A$ and of other pollutants $L$ that cause coronary heart disease $Y$.

In every item a variable ($L$ or $U$) causes both treatment and outcome,
so each is confounding in the sense of @def-confounding.
:::

::: {.notes}
**Terminology in the examples:**

- "Channeling" is often reserved for bias created by patient-specific risk factors $L$ that encourage doctors to prescribe a particular drug $A$ within a class of drugs.
- When subclinical disease $U$ causes both lack of exercise $A$ and clinical disease $Y$, and $L$ is unknown, this form of confounding is often called **reverse causation**.
- "Population stratification" is often reserved for bias from studying a mixture of individuals of different ethnic groups, so $U$ can stand for ethnicity.
:::

::: {.callout-note title="Notation: Bidirected Edges"}
Some authors draw an unmeasured common cause $U$ and its two arrows as a single bidirected edge between the two measured variables that $U$ causes.
This chapter keeps $U$ on the graph, in gray.
:::

---

::: {#rem-common-cause-associations-real}
## Associations from Common Causes Are Real

Early statistical writers, Yule (1903) for discrete variables and Pearson et al. (1899) for continuous ones,
called associations produced by common causes "fictitious", "illusory", "apparent", or "spurious" [@hernan2020causal, p. 92].
Such associations are real:
they would replicate in any large sample from the same population.
What they cannot be is an estimate of the effect of treatment.
:::

::: {.callout-important title="Standing Assumptions in This Chapter"}
Unless stated otherwise, this chapter assumes that

- positivity and consistency hold;
- every node in a causal diagram is measured without error;
- there are no selection nodes, so the data are a random sample from the population of interest;
- there is no random variability (the population is infinitely large).

Chapters 8, 9, and 10 relax these assumptions in turn [@hernan2020causal, pp. 92-93].
:::

::: {.notes}
The book reserves the term *confounding* for the common-cause structure of @def-confounding and uses other names for biases with other structures.
:::

## 7.2 Confounding and Exchangeability (pp. 93-95)

---

Exchangeability was defined in Chapter 2:
**marginal exchangeability** $Y^a \ind A$ ([Chapter 2](02-randomized-experiments.qmd#def-exchangeability))
and **conditional exchangeability** $Y^a \ind A \mid L$ ([Chapter 2](02-randomized-experiments.qmd#def-conditional-exchangeability)).
The next three results show what each one identifies.

::: {#prp-marginal-identification-ch7}
## Identification Under Marginal Exchangeability

Let $A$ be a binary treatment and $Y$ an outcome.
Assume marginal exchangeability $Y^a \ind A$,
positivity $\Pr[A = a] > 0$,
and consistency ($Y = Y^a$ whenever $A = a$), for $a = 0, 1$.
Then $\E{Y^a} = \E{Y \mid A = a}$ for $a = 0, 1$,
so the average causal effect equals the associational difference:

$$
\E{Y^{a=1}} - \E{Y^{a=0}} = \E{Y \mid A = 1} - \E{Y \mid A = 0}
$$ {#eq-marginal-identification-ch7}
:::

::: {.proof}
For each $a$,

\begin{align}
\E{Y^a}
&= \E{Y^a \mid A = a}
&& \text{(exchangeability; positivity)} \\
&= \E{Y \mid A = a}
&& \text{(consistency)}
\end{align}

Positivity makes the conditional mean in the first line well defined.
:::

::: {#exm-coin-flip-trial}
## A Marginally Randomized Experiment

In a trial in which a coin decides who is treated, nothing causes both $A$ and $Y$,
so $Y^a \ind A$ holds by design.
If 30% of the treated and 40% of the untreated die,
@prp-marginal-identification-ch7 gives a causal risk difference of $0.30 - 0.40 = -0.10$.
:::

---

::: {#prp-conditional-identification-ch7}
## Stratum-Specific Identification Under Conditional Exchangeability

Let $A$ be a treatment, $Y$ an outcome, and $L$ a discrete set of covariates.
Fix a treatment value $a$, and assume conditional exchangeability $Y^a \ind A \mid L$,
positivity $\Pr[A = a \mid L = l] > 0$ for every $l$ with $\Pr[L = l] > 0$,
and consistency.
Then, for every such $l$,

$$
\E{Y^a \mid L = l} = \E{Y \mid L = l, A = a}
$$ {#eq-conditional-identification-ch7}

so the conditional effects $\E{Y^{a=1} \mid L = l} - \E{Y^{a=0} \mid L = l}$ are identified by stratification.
:::

::: {.proof}
\begin{align}
\E{Y^a \mid L = l}
&= \E{Y^a \mid L = l, A = a}
&& \text{(conditional exchangeability; positivity)} \\
&= \E{Y \mid L = l, A = a}
&& \text{(consistency)}
\end{align}
:::

::: {#thm-standardization-ch7}
## Standardization Under Conditional Exchangeability

Under conditional exchangeability $Y^a \ind A \mid L$, positivity, and consistency,

$$\E{Y^a} = \sum_l \E{Y \mid L = l, A = a} \Pr[L = l]$$
:::

::: {.proof}
\begin{align}
\E{Y^a}
&= \sum_l \E{Y^a \mid L = l} \Pr[L = l]
&& \text{(law of total expectation)} \\
&= \sum_l \E{Y \mid L = l, A = a} \Pr[L = l]
&& \text{(@prp-conditional-identification-ch7)}
\end{align}

Here $L$ is discrete, the sums run over the $l$ with $\Pr[L = l] > 0$,
and positivity, $\Pr[A = a \mid L = l] > 0$ for each such $l$, ensures that the conditional means are defined.
:::

### Supplement: A Numerical Example

---

::: {#exm-standardization-numerical-ch7}
## Standardization Removes a Confounded Difference

This hypothetical example is not from the book.
Twenty individuals receive a drug $A$ or not;
$L = 1$ indicates a high-risk group;
$Y = 1$ indicates death.

| $L$ | $A$ | $n$ | Deaths | $\Pr[Y = 1 \mid A, L]$ |
|:---:|:---:|:---:|:------:|:----------------------:|
| 1 | 1 | 8 | 4 | 0.50 |
| 1 | 0 | 2 | 1 | 0.50 |
| 0 | 1 | 2 | 0 | 0.00 |
| 0 | 0 | 8 | 0 | 0.00 |

Crude risks:
$\Pr[Y = 1 \mid A = 1] = (4 + 0) / (8 + 2) = 0.40$ and
$\Pr[Y = 1 \mid A = 0] = (1 + 0) / (2 + 8) = 0.10$.

Assume $L$ is the only common cause of $A$ and $Y$, so that $Y^a \ind A \mid L$.
With $\Pr[L = 1] = 10/20 = 0.50$, standardization (@thm-standardization-ch7) gives

\begin{align}
\Pr[Y^{a=1} = 1] &= 0.50 \times 0.50 + 0.00 \times 0.50 = 0.25 \\
\Pr[Y^{a=0} = 1] &= 0.50 \times 0.50 + 0.00 \times 0.50 = 0.25
\end{align}

so the standardized risk difference is $0$, while the crude risk difference is $0.40 - 0.10 = 0.30$.
:::

::: {.notes}
The high-risk group ($L = 1$) is treated more often ($8/10$ versus $2/10$) and dies more often ($0.50$ versus $0.00$).
Within each level of $L$ the treated and untreated have the same risk, so the whole crude risk difference of $0.30$ is confounding.
:::

### The Backdoor Criterion

---

If we know the true causal DAG, how can we tell whether some set $L$ yields conditional exchangeability?
The book gives two approaches: the **backdoor criterion** (Pearl 1995) and the transformation of the DAG into a **SWIG** (Section 7.5).
The criterion uses *blocking*: conditioning on a non-collider on a path blocks it, and so does leaving a collider (and its descendants) unconditioned
(see [d-separation in Chapter 6](06-graphical-representation.qmd#from-d-separation-to-independence)).

::: {#def-backdoor-criterion-ch7}
## Backdoor Criterion

A set of covariates $L$ satisfies the **backdoor criterion** if

1. all backdoor paths between $A$ and $Y$ are blocked by conditioning on $L$, and
2. $L$ contains no variables that are descendants of treatment $A$.
:::

::: {#exm-backdoor-criterion-7-1}
## Checking the Backdoor Criterion in Figure 7.1

In Figure 7.1 the only backdoor path is $A \leftarrow L \rightarrow Y$.

- The set $\{L\}$ satisfies the criterion:
  $L$ is a non-collider on that path, so conditioning on it blocks the path, and $L$ is not a descendant of $A$.
- The empty set does not:
  with nothing conditioned on, the backdoor path stays open.
:::

---

::: {#thm-backdoor-exchangeability-ch7}
## Backdoor Criterion and Conditional Exchangeability

Let $\mathcal{G}$ be a causal DAG for treatment $A$, outcome $Y$, and covariates $L$,
interpreted as an FFRCISTG model ([Chapter 6](06-graphical-representation.qmd#graphs-and-counterfactuals)).

1. If $L$ satisfies the backdoor criterion (@def-backdoor-criterion-ch7), then $Y^a \ind A \mid L$ for every $a$.
2. If, in addition, the joint distribution is faithful to $\mathcal{G}$,
   then $Y^a \ind A \mid L$ for every $a$ only if $L$ satisfies the backdoor criterion.

Hence, under faithfulness, $Y^a \ind A \mid L$ holds **if and only if** $L$ satisfies the backdoor criterion [@hernan2020causal, pp. 93-94].
:::

Section 7.5 gives a short graphical argument for @thm-backdoor-exchangeability-ch7 once SWIGs are defined.
Checking every subset of measured non-descendants of $A$ against the criterion therefore tells us whether any of them achieves conditional exchangeability.

::: {.callout-note title="Technical Point 7.1: Does Conditional Exchangeability Imply the Backdoor Criterion?"}
Part 1 of @thm-backdoor-exchangeability-ch7 needs no faithfulness.
Part 2, the converse, depends on the counterfactual model:
it holds under an FFRCISTG model (Technical Point 6.3) but can fail under an NPSEM-IE model (@exm-npsem-ie-backdoor below).
The NPSEM-IE assumes cross-world independencies between counterfactuals,
which no randomized experiment could ever verify;
for that reason Robins did not assume them in the FFRCISTG model.
The book assumes an FFRCISTG model and faithfulness unless stated otherwise [@hernan2020causal, p. 94].
:::

::: {#exm-npsem-ie-backdoor}
## Conditional Exchangeability Without the Backdoor Criterion

Take the causal DAG with nodes $A$, $L$, $Y$ and arrows $A \rightarrow L$ and $A \rightarrow Y$.
The set $\{L\}$ fails the backdoor criterion, because $L$ is a descendant of $A$.

- Under an NPSEM-IE, $Y^a$ depends only on $a$ and an error term that is independent of the errors generating $A$ and $L$,
  so $Y^a \ind (A, L)$ and hence $Y^a \ind A \mid L$.
- Under an FFRCISTG model, conditioning on $L = l$ among those with $A = a'$ means conditioning on $L^{a'} = l$.
  For $a' \neq a$, independence of $Y^a$ and $L^{a'}$ is a cross-world statement that the model does not impose,
  so $Y^a \ind A \mid L$ cannot be assumed.
:::

### Two Settings in Which the Backdoor Criterion Holds

---

::: {#rem-backdoor-two-settings}
## Two Settings in Which the Backdoor Criterion Holds

1. **No common causes of treatment and outcome** (e.g., Figure 6.2):
   there are no backdoor paths, the empty set satisfies the criterion, and there is **no confounding**.
   This is a marginally randomized experiment;
   under faithfulness, marginal exchangeability is equivalent to no common causes of treatment and outcome.
2. **Common causes, but a set $L$ of measured non-descendants of $A$ blocks all backdoor paths** (e.g., Figure 7.1):
   there is confounding, but no confounding is left once $L$ is adjusted for.
   This is a conditionally randomized experiment, with an arrow $L \rightarrow A$ by design.
:::

::: {#def-sufficient-set}
## Sufficient Set for Confounding Adjustment

A set $L$ of measured non-descendants of $A$ is a **sufficient set for confounding adjustment** when conditioning on $L$ blocks all backdoor paths, that is, when the treated and the untreated are exchangeable within levels of $L$.
:::

::: {#def-no-unmeasured-confounding}
## No Unmeasured Confounding

There is **no unmeasured confounding** (given $L$) when the measured variables include a sufficient set $L$ for confounding adjustment (@def-sufficient-set),
so that no backdoor path needs an unmeasured variable to be blocked.
There may still be confounding:
the point is that $L$ is enough to remove it.
:::

::: {#exm-heart-transplant-sufficient-set}
## The Heart Transplant Study

In the heart transplant study of Chapter 2, the treated had a higher frequency of severe heart disease $L$, a common cause of transplant $A$ and death $Y$.
Had the treated not received a transplant, their risk of death would have exceeded the risk of the untreated,
so the two groups are not marginally exchangeable.
Within levels of $L$ they are exchangeable:
$\{L\}$ is a sufficient set, and there is no unmeasured confounding given $L$.
This second setting of @rem-backdoor-two-settings is also what investigators hope for in observational studies that measure many variables $L$.
:::

::: {.callout-warning title="The Backdoor Criterion Is Silent on Size and Sign"}
The criterion says whether bias is present, not how large it is or which way it goes.
An open backdoor path may carry only a weak association,
and two strong open paths may push the estimate in opposite directions and partly cancel.
Unmeasured confounding comes in degrees,
so investigators should think about its likely direction and magnitude (Fine Point 7.1).
:::

::: {.notes}
The book refers to Greenland and Robins (1986, 2009) for a detailed discussion of the relations between confounding and exchangeability.
:::

## 7.3 Confounding and the Backdoor Criterion (pp. 95-98)

---

::: {#exm-backdoor-figures-7-1-to-7-3}
## The Backdoor Criterion in Figures 7.1-7.3

- **Figure 7.1**: the backdoor path $A \leftarrow L \rightarrow Y$ is blocked by conditioning on $L$.
- **Figure 7.2**: the backdoor path $A \leftarrow L \leftarrow U \rightarrow Y$ could be blocked by $U$, which is unmeasured, but it is also blocked by $L$.
- **Figure 7.3**: the backdoor path $A \leftarrow U \rightarrow L \rightarrow Y$ is also blocked by $L$.

In all three there is confounding, but $\{L\}$ is a sufficient set, so there is **no unmeasured confounding given $L$** (@def-no-unmeasured-confounding).
:::

### M-bias: Figure 7.4

---

```{r}
#| label: dag-ch7-4-5
#| echo: false
#| fig-width: 8
#| fig-height: 3
#| fig-cap: "Left: Figure 7.4, in which L is a collider on the path A <- U2 -> L <- U1 -> Y. Right: Figure 7.5, which adds the arrow L -> A."
dag_7_4 <- dagitty('dag {
  U2 [pos="0,2"]; U1 [pos="2,2"]; L [pos="1,1"]; A [pos="0,0"]; Y [pos="2,0"]
  U2 -> A; U2 -> L; U1 -> L; U1 -> Y; A -> Y
}')
dag_7_5 <- dagitty('dag {
  U2 [pos="0,2"]; U1 [pos="2,2"]; L [pos="1,1"]; A [pos="0,0"]; Y [pos="2,0"]
  U2 -> A; U2 -> L; U1 -> L; U1 -> Y; L -> A; A -> Y
}')
draw_dag(dag_7_4, unmeasured = c("U1", "U2"), title = "Figure 7.4") +
  draw_dag(dag_7_5, unmeasured = c("U1", "U2"), title = "Figure 7.5")
```

::: {#exm-figure-7-4-no-confounding}
## No Confounding in Figure 7.4

In Figure 7.4 $A$ and $Y$ share no common cause, so there is **no confounding**:
the backdoor path $A \leftarrow U_2 \rightarrow L \leftarrow U_1 \rightarrow Y$ is blocked by the collider $L$,
and, by @prp-marginal-identification-ch7, $\Pr[Y^a = 1] = \Pr[Y = 1 \mid A = a]$.
:::

::: {#exm-pap-smear}
## Physical Activity and Cervical Cancer

Let $A$ be physical activity, $Y$ cervical cancer, $U_1$ a pre-cancer lesion,
$L$ a Pap smear (a diagnostic test for pre-cancer),
and $U_2$ a health-conscious personality that leads to more physical activity and more doctor visits.
Under Figure 7.4 the effect of $A$ on $Y$ is unconfounded,
and there is no need to adjust for $L$ to compute $\Pr[Y^{a=1} = 1]$ or $\Pr[Y^{a=0} = 1]$ [@hernan2020causal, p. 95].
:::

---

::: {.callout-warning title="Adjusting for a Collider Creates Bias"}
In Figure 7.4, conditioning on the collider $L$ opens the path $A \leftarrow U_2 \rightarrow L \leftarrow U_1 \rightarrow Y$,
so adjusting for $L$ (by stratification or by standardization over $L$) **creates** bias where there was none.
:::

::: {#def-m-bias-ch7}
## M-bias

**M-bias** is the bias that arises from conditioning on a variable $L$ that is a common effect of two variables $U_1$ and $U_2$,
where $U_2$ is a cause of (or otherwise associated with) treatment and $U_1$ is a cause of (or otherwise associated with) the outcome,
and $U_1$ and $U_2$ are marginally independent.
The book classifies it as a form of **selection bias**, the subject of Chapter 8 [@hernan2020causal, p. 97].
:::

::: {#exm-m-bias-pap-smear}
## M-bias in the Pap Smear Example

In @exm-pap-smear, comparing physical activity and cervical cancer only among women with the same Pap smear result conditions on $L$,
a common effect of the pre-cancer lesion $U_1$ and of health-conscious personality $U_2$.
Within a stratum of $L$, physical activity and cervical cancer generally become associated through $U_2$ and $U_1$,
so the stratum-specific association mixes the causal effect with bias, even though the unadjusted association is causal.
:::

::: {#rem-unconditional-without-conditional}
## Unconditional Without Conditional Exchangeability

Figure 7.4 is the first example in which $Y^a \ind A$ holds but $Y^a \ind A \mid L$ does not.
The average causal effect is identified by @prp-marginal-identification-ch7,
but the effects within levels of $L$ generally are not,
and standardizing over $L$ gives a biased estimate of $\Pr[Y^a = 1]$.
:::

::: {.notes}
**Why "M-bias":**
the structure was described by Greenland, Pearl, and Robins (1999) and named M-bias by Greenland (2003),
because $U_2$, $L$, $U_1$ resemble an M lying on its side [@hernan2020causal, p. 97].
That unconditional effects can be identified while conditional effects are not was shown non-graphically by Greenland and Robins (1986).

If $U_1$ caused $U_2$, $U_2$ caused $U_1$, or an unmeasured $U_3$ caused both,
$A$ and $Y$ would have a common cause, and there would be neither unconditional nor conditional exchangeability given $L$.
:::

### Intractable Bias: Figure 7.5

---

::: {#exm-intractable-bias-7-5}
## Intractable Bias in Figure 7.5

Figure 7.5 adds the arrow $L \rightarrow A$, which creates the open backdoor path $A \leftarrow L \leftarrow U_1 \rightarrow Y$: there is confounding.

- Conditioning on $L$ blocks $A \leftarrow L \leftarrow U_1 \rightarrow Y$,
- but opens $A \leftarrow U_2 \rightarrow L \leftarrow U_1 \rightarrow Y$, on which $L$ is a collider.

So neither the empty set nor $\{L\}$ satisfies the backdoor criterion:
there is neither unconditional exchangeability nor conditional exchangeability given $L$.
:::

::: {.callout-note title="Being a Collider Is Path-Specific"}
In Figure 7.5, $L$ is a collider on $A \leftarrow U_2 \rightarrow L \leftarrow U_1 \rightarrow Y$
but a non-collider on $A \leftarrow L \leftarrow U_1 \rightarrow Y$.
A variable is a collider only relative to a given path.
:::

::: {#rem-escaping-figure-7-5}
## Measuring One More Variable

The bias in Figure 7.5 disappears if one more variable is measured:

- a variable $L_1$ that sits between $U_1$ and either $A$ or $Y$ gives conditional exchangeability given $L_1$;
- a variable $L_2$ that sits between $U_2$ and either $A$ or $L$ gives conditional exchangeability given $\{L_2, L\}$.

**Figure 7.6** (not drawn here) is Figure 7.5 with the arrow $U_1 \rightarrow Y$ replaced by $U_1 \rightarrow L_1 \rightarrow Y$
and the arrow $U_2 \rightarrow A$ replaced by $U_2 \rightarrow L_2 \rightarrow A$.
:::

---

::: {.callout-note title="Fine Point 7.1: The Strength and Direction of Confounding Bias"}
Knowing that confounding may be present is only half the story:
investigators also want to know whether the unadjusted estimate is too large or too small, and by how much.
Signed causal diagrams (@def-signed-causal-diagram) give a rule of thumb for the direction,
@exm-signed-smoking-transplant applies it,
and @rem-confounding-magnitude covers the magnitude [@hernan2020causal, p. 96].
:::

::: {#def-signed-causal-diagram}
## Signed Causal Diagram; Positive and Negative Confounding

Take Figure 7.1 with dichotomous $L$, $A$, and $Y$.
A **signed causal diagram** puts a $+$ on the arrow $L \rightarrow A$ if $L$ has a positive average causal effect on $A$
($\Pr[A = 1]$ is higher when $L = 1$ than when $L = 0$) and a $-$ if the effect is negative,
and signs the arrow $L \rightarrow Y$ in the same way.
The confounding is **positive** if the two signs agree and **negative** if they differ.
In this simple setting, positive confounding biases the unadjusted estimate upward and negative confounding biases it downward.
:::

::: {#exm-signed-smoking-transplant}
## Smoking and Heart Transplant

Suppose a study of heart transplant $A$ on death $Y$ finds a risk ratio of 0.6, and a critic suspects confounding by smoking $L$.
Smokers are less likely to receive a transplant ($L \rightarrow A$ is $-$) and more likely to die ($L \rightarrow Y$ is $+$),
so by @def-signed-causal-diagram the confounding is negative and the unadjusted risk ratio is too low.
Because the transplant group has fewer smokers, it would have lower mortality even if transplant did nothing,
so adjusting for smoking moves the estimate upward (the book's illustration: from 0.6 to 0.7).
Failing to adjust exaggerates the apparent benefit.
:::

::: {.callout-warning title="The Sign Rule Can Fail"}
The sign rule of @def-signed-causal-diagram may fail in more complex diagrams or with non-dichotomous variables (VanderWeele, Hernán, and Robins 2008).
:::

::: {#rem-confounding-magnitude}
## The Magnitude of Confounding Bias

A large confounding bias needs both a strong association between confounder and treatment and a strong association between confounder and outcome (conditional on treatment);
for a discrete confounder, the bias also depends on how common it is.
When the confounders are unknown, sensitivity analyses repeat the analysis under a range of assumed bias sizes,
which turns educated guesses into a plausible range for the bias [@hernan2020causal, p. 96].
:::

### Confounders and Non-confounders

---

Suppose data on $L$, $A$, and $Y$ suffice to identify the causal effect, as in Figures 7.1-7.4.

::: {#def-confounder-ch7}
## Confounder (Given Data on L, A, Y)

$L$ is a **confounder** if data on $A$ and $Y$ alone do not suffice for identification,
that is, if there is conditional exchangeability given $L$ but not unconditional exchangeability (structural confounding).
$L$ is a **non-confounder** if data on $A$ and $Y$ alone suffice, that is, if there is unconditional exchangeability [@hernan2020causal, p. 96].
:::

::: {#exm-confounders-figures-7-1-to-7-4}
## Confounders and Non-confounders in Figures 7.1-7.4

- In Figures 7.1-7.3, $L$ is a confounder:
  $\Pr[Y^a = 1]$ is identified by the standardized risk $\sum_l \Pr[Y = 1 \mid A = a, L = l] \Pr[L = l]$ (@thm-standardization-ch7).
  In Figures 7.2 and 7.3, $L$ is not a common cause of $A$ and $Y$, yet it is a confounder because it is needed to block the backdoor path through $U$.
- In Figure 7.4, $L$ is a non-confounder, and $\Pr[Y^a = 1] = \Pr[Y = 1 \mid A = a]$.
  Standardizing by $L$ would be biased.
:::

::: {.notes}
An informal definition for Figures 7.1 to 7.4 is "a confounder is any variable that can be used to adjust for confounding."
This definition is not circular, because confounding was defined first,
just as "a musician is a person who plays music" is not circular once music has been defined [@hernan2020causal, p. 96].
:::

### Two Definitions of Confounding

---

The causal diagrams of this section show two structural sources of lack of exchangeability through open backdoor paths:

- common causes of treatment and outcome, which the book calls **confounding** (@def-confounding);
- conditioning on a common effect, which the book calls **selection bias** (as in @def-m-bias-ch7).

::: {#def-confounding-backdoor}
## Confounding as Any Open-Backdoor-Path Bias

An alternative structural definition calls **confounding** any bias due to an open backdoor path between $A$ and $Y$,
whether the path is open because of a common cause or because of conditioning on a collider.
Equivalently, confounding is any systematic bias that randomized assignment of $A$ would eliminate.
:::

::: {#exm-confounding-backdoor-figure-7-4}
## Figure 7.4 Under the Two Definitions

In Figure 7.4, the bias from adjusting for $L$ is selection bias under @def-confounding but confounding under @def-confounding-backdoor.
Randomizing $A$ would remove it:
random assignment rules out an unmeasured common cause $U_2$ of $A$ and $L$,
so conditioning on $L$ would no longer open a backdoor path.
Every other bias in Figures 7.1-7.5 gets the same label under both definitions.
:::

::: {#rem-two-definitions-taste}
## The Choice of Definition Has No Practical Consequences

Under @def-confounding, whether confounding exists is a fact about the population, whatever the analysis.
Under @def-confounding-backdoor it depends on the analysis:
in Figure 7.4 there is no confounding if we do not adjust for $L$, but there is if we do.
Either way, what can be identified depends only on whether conditional or unconditional exchangeability holds,
so the choice is a matter of taste [@hernan2020causal, p. 98].
:::

---

::: {.callout-note title="Fine Point 7.2: Identification of Conditional and Unconditional Effects"}
Which effects can be identified depends on which variables are measured besides $A$ and $Y$.
@exm-figure-7-6-identification works this out for Figure 7.6 (@rem-escaping-figure-7-5) [@hernan2020causal, p. 98].
:::

::: {#exm-figure-7-6-identification}
## What Can Be Identified in Figure 7.6

- Measuring only $L_2$: no exchangeability given $L_2$;
  no causal effects are identified.
- Measuring $L_2$ and $L$: conditional exchangeability given $\{L_2, L\}$ (but not given either alone).
  Identified are:
  - effects within joint strata of $L$ and $L_2$, via $\E{Y \mid A = a, L = l, L_2 = l_2}$;
  - the unconditional effect, via $\sum_{l, l_2} \E{Y \mid A = a, L = l, L_2 = l_2} \Pr[L = l, L_2 = l_2]$;
  - effects within strata of $L$, via $\sum_{l_2} \E{Y \mid A = a, L = l, L_2 = l_2} \Pr[L_2 = l_2 \mid L = l]$;
  - effects within strata of $L_2$, via $\sum_{l} \E{Y \mid A = a, L = l, L_2 = l_2} \Pr[L = l \mid L_2 = l_2]$.
- Measuring only $L_1$: effects within strata of $L_1$ and the unconditional effect.
- Measuring $L_1$ and $L$: additionally, effects within joint strata of $L_1$ and $L$, and within strata of $L$.
- Measuring $L$, $L_1$, and $L_2$: additionally, effects within joint strata of all three.

Each formula identifies a counterfactual mean $\E{Y^a \mid \cdot}$ under positivity and consistency.
:::

## 7.4 Confounding and Confounders (pp. 98-101)

---

The structural approach needs prior knowledge of the causal DAG, including all shared causes (measured or not) of $A$ and $Y$;
the backdoor criterion then says what to adjust for.
The **traditional approach** instead labels as confounders the variables meeting mostly associational conditions,
mandates adjusting for them, and declares confounding when adjusted and unadjusted estimates differ.

::: {#def-traditional-confounder}
## Traditional Definition of Confounder

A variable is a **confounder** under the traditional approach if it

1. is associated with the treatment,
2. is associated with the outcome conditional on the treatment (often replaced by "in the untreated"), and
3. does not lie on a causal pathway between treatment and outcome.
:::

::: {#exm-traditional-figures-7-1-to-7-4}
## Where the Traditional Approach Agrees and Disagrees

- **Figures 7.1-7.3**: $L$ meets all three conditions, matching the backdoor criterion.
- **Figure 7.4**: $L$ meets all three conditions
  (it shares $U_2$ with $A$ and $U_1$ with $Y$, and is not on the causal pathway),
  so the traditional approach says to adjust for it,
  although there is no confounding and adjusting causes selection bias.
- **Figure 7.7** (not drawn here) is a second example in which the traditional approach leads to harmful adjustment for $L$.
:::

::: {#rem-modified-traditional-definition}
## Patching Condition 2 Does Not Work

Replacing condition 2 of @def-traditional-confounder by the structural condition "it is a cause of the outcome" fixes Figure 7.4,
but then $L$ in Figure 7.2 would no longer count as a confounder, although it must be adjusted for.
Technical Point 7.2 gives a repair that works.
:::

::: {.callout-warning title="Associations Cannot Define Confounding"}
A definition of confounder built almost entirely on statistical associations can advise adjusting for a "confounder" when no structural confounding exists.
Nor does a change in estimate after adjustment prove confounding:
adjusted and unadjusted estimates can differ because adjustment for a non-confounder created selection bias (Chapter 8),
or because the effect measure is noncollapsible (Fine Point 4.3).
Definitions of confounding based on change in estimates were abandoned long ago for these reasons [@hernan2020causal, p. 100].
:::

::: {.notes}
Technically, investigators do not need full structural knowledge;
they only need to know a set of variables that guarantees conditional exchangeability.
:::

### Confounding Is Absolute; Confounder Is Relative

---

::: {#rem-confounder-relative}
## Confounding Is Absolute; Confounder Is Relative

The structural approach first identifies the sources of confounding (the common causes of treatment and outcome) and then a sufficient adjustment set.
Whether a variable belongs to a sufficient set depends on the other variables in it.
In Figures 7.2 and 7.3, $L$ is needed only because $U$ is unmeasured;
given $U$, $L$ would not be a confounder.
Given a causal DAG, **confounding is an absolute concept, whereas confounder is a relative one** [@hernan2020causal, p. 100].
:::

::: {#rem-structural-approach-advantages-ch7}
## Two Advantages of the Structural Approach

1. It prevents inconsistencies between beliefs and actions:
   if you believe Figure 7.4, you will not adjust for $L$, whatever a non-structural definition says.
2. It makes the researchers' assumptions explicit, so others can criticize them.
:::

::: {.notes}
VanderWeele and Shpitser (2013) also proposed a formal definition of confounder.
No approach guarantees that the researchers' DAG is correct,
so a chosen adjustment set may still fail to remove confounding or may introduce selection bias.
:::

---

```{r}
#| label: dag-ch7-8
#| echo: false
#| fig-width: 4
#| fig-height: 2.5
#| fig-cap: "Figure 7.8. L is a surrogate (proxy) for the unmeasured common cause U of A and Y."
dag_7_8 <- dagitty('dag {
  U [pos="1,1"]; L [pos="2,1"]; A [pos="0,0"]; Y [pos="2,0"]
  U -> L; U -> A; U -> Y; A -> Y
}')
draw_dag(dag_7_8, unmeasured = "U")
```

::: {.callout-note title="Fine Point 7.3: Surrogate Confounders"}
A measured variable can help with confounding without lying on any backdoor path.
@def-surrogate-confounder names such variables and @exm-income-ses gives an example [@hernan2020causal, p. 100].
:::

::: {#def-surrogate-confounder}
## Surrogate Confounder

A **surrogate confounder** is a measured variable $L$ that is not on a backdoor path between $A$ and $Y$
but is associated with an unmeasured confounder $U$,
so that adjusting for $L$ can reduce, though never completely remove, the confounding by $U$.
:::

::: {#exm-income-ses}
## Income as a Surrogate for Socioeconomic Status

In Figure 7.8 the unmeasured $U$ (e.g., socioeconomic status) confounds the effect of physical activity $A$ on cardiovascular disease $Y$,
and the measured $L$ (e.g., income) is a proxy for $U$.
$L$ is not on a backdoor path, but adjusting for it may remove some of the confounding by $U$;
if $L$ were perfectly correlated with $U$, conditioning on $L$ would be the same as conditioning on $U$.
If $L$ is a binary, nondifferentially misclassified version of $U$, conditioning on $L$ partially blocks $A \leftarrow U \rightarrow Y$ under some weak conditions
(Greenland 1980; Ogburn and VanderWeele 2012).
So one typically prefers to adjust for $L$.
:::

::: {.callout-tip title="Collect Many Surrogates"}
One strategy against unmeasured confounding is to measure and adjust for as many surrogate confounders as possible (see Chapter 18).
:::

---

::: {.callout-note title="Technical Point 7.2: Fixing the Traditional Definition of Confounder"}
The traditional definition (@def-traditional-confounder) rests on two incorrect statistical criteria (conditions 1 and 2) and one incorrect causal criterion (condition 3).
@def-nonconfounder-given-data replaces them,
following Robins (1997, Theorem 4.3) and Greenland, Pearl, and Robins (1999),
@prp-nonconfounder-given-data shows what the replacement buys,
and @exm-nonconfounder-figure-7-4 applies it to Figure 7.4 [@hernan2020causal, p. 101].
:::

::: {#def-nonconfounder-given-data}
## Non-confounder Given Data on L

Let $L$ (measured) and $U$ (possibly unmeasured) be sets of non-descendants of $A$
with conditional exchangeability $Y^a \ind A \mid L, U$.
$U$ is a **non-confounder given data on $L$** if $U$ can be split into disjoint subsets $U_1$ and $U_2$ ($U = U_1 \cup U_2$, $U_1 \cap U_2 = \emptyset$) such that

1. $U_1$ and $A$ are not associated within strata of $L$: $U_1 \ind A \mid L$, and
2. $U_2$ and $Y$ are not associated within joint strata of $A$, $L$, and $U_1$: $U_2 \ind Y \mid A, L, U_1$.

$U_1$ and $U_2$ may be associated with each other.
:::

::: {#prp-nonconfounder-given-data}
## Adjusting for L Alone Suffices

Let $L$ and $U$ be discrete, and suppose $U$ is a non-confounder given data on $L$ (@def-nonconfounder-given-data).
Fix a treatment value $a$ and assume consistency and positivity, $\Pr[A = a \mid L = l, U = u] > 0$ for every $(l, u)$ with $\Pr[L = l, U = u] > 0$.
Then, for every $l$ with $\Pr[L = l] > 0$,

$$
\E{Y^a \mid L = l} = \E{Y \mid A = a, L = l}
$$ {#eq-nonconfounder-given-data}

so $\E{Y^a} = \sum_l \E{Y \mid A = a, L = l} \Pr[L = l]$:
standardization over $L$ alone identifies the counterfactual mean.
:::

::: {.proof}
This derivation is ours, not the book's.
Write $\mu(u_1) \eqdef \E{Y \mid A = a, L = l, U_1 = u_1}$.
By condition 2, $\E{Y \mid A = a, L = l, U_1 = u_1, U_2 = u_2} = \mu(u_1)$.
First, by the law of total expectation, conditional exchangeability, and consistency,

\begin{align}
\E{Y^a \mid L = l}
&= \sum_{u} \E{Y^a \mid L = l, U = u} \Pr[U = u \mid L = l] \\
&= \sum_{u} \E{Y \mid A = a, L = l, U = u} \Pr[U = u \mid L = l] \\
&= \sum_{u_1} \mu(u_1) \Pr[U_1 = u_1 \mid L = l]
\end{align}

Second, by the law of total expectation and condition 2,

\begin{align}
\E{Y \mid A = a, L = l}
&= \sum_{u} \E{Y \mid A = a, L = l, U = u} \Pr[U = u \mid A = a, L = l] \\
&= \sum_{u_1} \mu(u_1) \Pr[U_1 = u_1 \mid A = a, L = l] \\
&= \sum_{u_1} \mu(u_1) \Pr[U_1 = u_1 \mid L = l]
\end{align}

where the last line uses condition 1.
The two right-hand sides agree, which proves @eq-nonconfounder-given-data;
averaging over $\Pr[L = l]$ gives the formula for $\E{Y^a}$.
:::

::: {#exm-nonconfounder-figure-7-4}
## No Confounding in Figure 7.4, Without the Backdoor Criterion

In Figure 7.4 take $L = \emptyset$ (adjust for nothing) and $U = \{U_1, U_2\}$, so that $Y^a \ind A \mid U_1, U_2$.

1. $U_1 \ind A$: the only path from $U_1$ to $A$ runs through the collider $L$, which is not conditioned on.
2. $U_2 \ind Y \mid A, U_1$: the path $U_2 \rightarrow A \rightarrow Y$ is blocked by conditioning on $A$, and $U_2 \rightarrow L \leftarrow U_1 \rightarrow Y$ is blocked by $L$ and by $U_1$.

So $\{U_1, U_2\}$ is a non-confounder given no data, and by @prp-nonconfounder-given-data $\E{Y^a} = \E{Y \mid A = a}$:
there is no confounding, as found with the backdoor criterion in @exm-figure-7-4-no-confounding.
Here both independences are read off the DAG by d-separation, which implies independence under the causal Markov assumption.
:::

## 7.5 Single-World Intervention Graphs (pp. 101-102)

---

The equivalence between exchangeability and the backdoor criterion seems "rather magical" because counterfactuals do not appear on causal diagrams.
**Single-world intervention graphs (SWIGs)** put the counterfactual variables on the graph,
so that exchangeability can be read off directly by d-separation.

::: {#def-swig}
## Single-World Intervention Graph (SWIG)

A **SWIG** depicts the variables and causal relations that would be observed in a hypothetical world in which all individuals received treatment level $a$,
a counterfactual world created by a single intervention.
It is obtained from a causal DAG as follows:

- Split the treatment node into a left side $A$ (the **natural value** of treatment, the value that would have been observed without intervention),
  which keeps all arrows into $A$, and a right side $a$ (the intervention value), which inherits all arrows out of $A$.
  There is no arrow from $A$ to $a$, because $a$ is a constant.
- Replace each descendant of treatment by its counterfactual, e.g., $Y$ by $Y^a$.
  Non-descendants of $A$ keep their factual labels, because treatment does not affect them.
:::

::: {#exm-swig-7-9}
## SWIGs for Figures 7.2 and 7.4

- **Figure 7.9** is the SWIG of Figure 7.2:
  $U \rightarrow L \rightarrow A$, $U \rightarrow Y^a$, $a \rightarrow Y^a$.
  Every path between $Y^a$ and $A$ is blocked by conditioning on $L$, so $Y^a \ind A \mid L$.
- **Figure 7.10** is the SWIG of Figure 7.4:
  $U_2 \rightarrow A$, $U_2 \rightarrow L \leftarrow U_1 \rightarrow Y^a$, $a \rightarrow Y^a$.
  Without conditioning, the path through the collider $L$ is blocked, so $Y^a \ind A$.
  Conditioning on $L$ opens $Y^a \leftarrow U_1 \rightarrow L \leftarrow U_2 \rightarrow A$, so $Y^a \ind A \mid L$ fails.
:::

::: {.proof}
## Sketch of a proof of @thm-backdoor-exchangeability-ch7

On the SWIG, $Y^a$ is d-separated from $A$ given $L$ **if and only if** $L$ is a non-descendant of $A$ that blocks all backdoor paths from $A$ to $Y$ [@hernan2020causal, p. 102]:
the natural value $A$ keeps only the arrows into treatment, so every path between $A$ and $Y^a$ starts with an arrow into $A$, as a backdoor path does,
and a descendant of $A$ appears on the SWIG only as a counterfactual, so the factual variable cannot be in the conditioning set.
Under an FFRCISTG model, d-separation on the SWIG implies $Y^a \ind A \mid L$ (part 1);
under faithfulness, lack of d-separation implies dependence (part 2).
:::

::: {.callout-note title="Notation: Arrows Out of the Intervention Node"}
In the single-intervention world, $a$ is a constant and cannot affect other variables.
SWIGs still draw arrows from $a$, to keep track of which variables $A$ directly affects in the original DAG.
:::

::: {.notes}
**Origins:**
Richardson and Robins (2013) showed that SWIGs overcome some shortcomings of the twin causal diagrams previously proposed by Balke and Pearl (1994) [@hernan2020causal, p. 101].

**Is the natural value measurable?**
The book assumes the natural value $A$ is well defined even though it is generally not observed under intervention $a$.
It notes experiments suggesting that electroencephalogram recordings can detect a choice up to 1/2 second before people become conscious of it,
which in principle would allow measuring $A$ and still intervening.
:::

## 7.6 Confounding Adjustment (pp. 102-107)

---

Without randomization, causal inference relies on the uncheckable assumption that the measured $L$ is a sufficient set for confounding adjustment (@def-sufficient-set).
Under that assumption, methods that adjust for $L$ fall into two categories.

::: {#def-g-methods-ch7}
## G-methods and Conventional Stratification-Based Methods

Assume conditional exchangeability given $L$.

- **G-methods** ("g" for generalized) are standardization, IP weighting, and g-estimation.
  They use conditional exchangeability given $L$ to estimate the causal effect of $A$ on $Y$ in the whole population or in any subset of it.
- **Conventional stratification-based methods** are stratification (including restriction) and matching.
  They use conditional exchangeability given $L$ to estimate the association between $A$ and $Y$ within subsets defined by $L$.
:::

::: {#exm-heart-transplant-methods}
## Both Kinds of Method in the Heart Transplant Study

The heart transplant study adjusted for confounding by disease severity $L$ with g-methods in Sections 2.4 (standardization) and 2.5 (IP weighting),
and with conventional stratification-based methods in Sections 4.4 (stratification) and 4.5 (matching).
Part II covers their model-based extensions:
the parametric g-formula, IP weighting of marginal structural models, and g-estimation of structural nested models,
and, on the conventional side, outcome regression.
:::

::: {#rem-deleting-vs-conditioning}
## "Deleting" an Arrow Versus Conditioning

Standardization and IP weighting simulate the $A$-$Y$ association in the population if the backdoor paths through $L$ did not exist;
IP weighting, for example, creates a pseudo-population in which $A$ is independent of $L$, "deleting" the arrow $L \rightarrow A$.
Stratification does not delete that arrow but computes the effect in a subset, represented by a selection box.
Part III explains why deleting the arrow is advantageous with time-varying treatments,
why g-estimation is the only generally valid stratification-based method,
and why conventional stratification-based methods can cause selection bias with time-varying confounders (Chapter 20).
:::

::: {.notes}
A common variation of stratification and matching replaces $L$ by the estimated probability of treatment $\Pr[A = 1 \mid L]$, the **propensity score** (Rosenbaum and Rubin 1983;
Chapter 15).
:::

---

::: {.callout-note title="Fine Point 7.4: Confounders Cannot Be Descendants of Treatment, but Can Be in the Future of Treatment"}
The backdoor criterion excludes descendants of $A$, not variables measured after $A$.
@exm-descendant-blocks-backdoor shows why descendants are excluded,
and @rem-topology-not-time shows that timing alone does not matter [@hernan2020causal, p. 103].
:::

::: {#exm-descendant-blocks-backdoor}
## A Descendant That Blocks Every Backdoor Path (Figure 7.11)

Figure 7.11 (not drawn here) has arrows $U \rightarrow A$, $U \rightarrow L$, $A \rightarrow L$, and $L \rightarrow Y$, with $U$ unmeasured.
$L$ is a descendant of $A$ that blocks the only backdoor path, $A \leftarrow U \rightarrow L \rightarrow Y$, and the effect of $A$ on $Y$ is entirely through $L$.
Conditioning on $L$ opens no collider path, but it blocks the causal pathway, so adjusting for $L$ does not remove bias.
Hence conditional exchangeability $Y^a \ind A \mid L$ must fail.
On the SWIG (Figure 7.12: $U \rightarrow A$, $U \rightarrow L^a$, $a \rightarrow L^a \rightarrow Y^a$) $L$ is replaced by the counterfactual $L^a$, and we can read off $Y^a \ind A \mid L^a$ but not $Y^a \ind A \mid L$, since $L$ is not on the graph.
(Under an FFRCISTG model, an independence that cannot be read off the SWIG cannot be assumed to hold.)
:::

::: {#rem-topology-not-time}
## Topology, Not Timing

The problem in @exm-descendant-blocks-backdoor is that $L$ is a **descendant** of $A$, not that $L$ occurs **after** $A$.
Without the arrow $A \rightarrow L$, $L$ would be a non-descendant that blocks all backdoor paths, and adjusting for it would remove all bias even if $L$ were measured after $A$.
What matters is the topology of the causal diagram, not the time order of the nodes.
:::

### Methods That Do Not Require Conditional Exchangeability

---

Some methods handle confounding without conditional exchangeability:

- difference-in-differences and negative outcome controls (Technical Point 7.3);
- proximal inference (Technical Point 7.3);
- the front door criterion (Technical Point 7.4);
- instrumental variable estimation (Chapter 16).

::: {.callout-warning title="Other Methods Need Other Unverifiable Assumptions"}
These methods replace conditional exchangeability with other assumptions that are just as unverifiable,
so the choice of method depends on which unverifiable assumptions are more plausible in a given setting.
:::

### Expert Knowledge and the Critic

---

::: {.callout-tip title="Use Expert Knowledge of the Causal Structure"}
Conditional exchangeability may be unrealistic, but expert knowledge about the causal structure helps get close to it:

- measure many non-descendants $L$ of treatment in the hope of blocking all backdoor paths;
- avoid adjusting for variables affected by treatment or by the outcome;
- when several causal structures are plausible, analyze under each and state the assumptions each requires.
:::

::: {#exm-scientific-critique}
## A Logical Versus a Scientific Criticism

A critic who says only "your observational study may be confounded" makes a logical, not a scientific, statement:
it is true of every observational study.
A scientific criticism names a source, such as "confounding due to cigarette smoking, a common cause through which a backdoor path may remain open".
That gives a testable challenge: adjust for smoking or, if smoking was not measured, conduct a sensitivity analysis [@hernan2020causal, pp. 104-105].
:::

::: {.notes}
Hernán et al. (2002) give a practical example of using expert knowledge of the causal structure to evaluate confounding.
One can never be certain that the causal structures considered include the true one;
this uncertainty is unavoidable with observational data.

Valid causal inference from observational data also requires the absence of selection and measurement biases,
which, unlike confounding, can arise in randomized experiments too.
Chapter 8 turns to selection bias.
:::

---

```{r}
#| label: dag-ch7-13-14
#| echo: false
#| fig-width: 8
#| fig-height: 3
#| fig-cap: "Left: Figure 7.13, with negative outcome control C (the pre-treatment outcome). Right: Figure 7.14, in which M fully mediates the effect of A on Y. U is unmeasured."
dag_7_13 <- dagitty('dag {
  U [pos="1,1"]; C [pos="0,1"]; A [pos="0,0"]; Y [pos="2,0"]
  U -> C; U -> A; U -> Y; C -> Y; A -> Y
}')
dag_7_14 <- dagitty('dag {
  U [pos="1,1"]; A [pos="0,0"]; M [pos="1,0"]; Y [pos="2,0"]
  U -> A; U -> Y; A -> M; M -> Y
}')
draw_dag(dag_7_13, unmeasured = "U", title = "Figure 7.13") +
  draw_dag(dag_7_14, unmeasured = "U", title = "Figure 7.14")
```

::: {.callout-note title="Technical Point 7.3: Difference-in-Differences and Negative Outcome Controls"}
A variable that treatment cannot affect but that shares the unmeasured confounders of the outcome can measure the confounding.
@def-negative-outcome-control and @def-additive-equi-confounding set up the method,
@prp-difference-in-differences gives the identification result,
@exm-aspirin-blood-pressure illustrates it,
and @rem-proximal-inference points to a more general approach [@hernan2020causal, p. 106].
:::

::: {#def-negative-outcome-control}
## Negative Outcome Control

A **negative outcome control** for the effect of $A$ on $Y$ is a variable $C$ that $A$ cannot cause but that shares the unmeasured causes $U$ of $A$ and $Y$ (Figure 7.13).
The outcome measured just before treatment is the usual example.
Because $A$ has no effect on $C$, $\E{C \mid A = 1} - \E{C \mid A = 0}$ measures the additive confounding for the effect of $A$ on $C$.
:::

::: {#def-additive-equi-confounding}
## Additive Equi-confounding

A binary treatment $A$, outcome $Y$, and negative outcome control $C$ satisfy **additive equi-confounding** if the additive confounding for $Y$ under no treatment equals the additive confounding for $C$:

$$
\E{Y^0 \mid A = 1} - \E{Y^0 \mid A = 0} = \E{C \mid A = 1} - \E{C \mid A = 0}
$$ {#eq-additive-equi-confounding}
:::

::: {#prp-difference-in-differences}
## Difference-in-Differences

Let $A$ be a binary treatment with $0 < \Pr[A = 1] < 1$, $Y$ an outcome, and $C$ a negative outcome control (@def-negative-outcome-control) on the same scale as $Y$.
Assume consistency and additive equi-confounding (@eq-additive-equi-confounding).
Then the average causal effect in the treated, $\ATT \eqdef \E{Y^1 - Y^0 \mid A = 1}$, is identified:

$$
\ATT = \paren{\E{Y \mid A = 1} - \E{Y \mid A = 0}} - \paren{\E{C \mid A = 1} - \E{C \mid A = 0}}
$$ {#eq-difference-in-differences}
:::

::: {.proof}
\begin{align}
\E{Y^1 - Y^0 \mid A = 1}
&= \E{Y^1 \mid A = 1} - \E{Y^0 \mid A = 1}
&& \text{(linearity)} \\
&= \E{Y \mid A = 1} - \E{Y^0 \mid A = 1}
&& \text{(consistency)} \\
&= \E{Y \mid A = 1} - \E{Y^0 \mid A = 0} - \paren{\E{Y^0 \mid A = 1} - \E{Y^0 \mid A = 0}}
&& \text{(add and subtract } \E{Y^0 \mid A = 0}\text{)} \\
&= \E{Y \mid A = 1} - \E{Y \mid A = 0} - \paren{\E{Y^0 \mid A = 1} - \E{Y^0 \mid A = 0}}
&& \text{(consistency)} \\
&= \paren{\E{Y \mid A = 1} - \E{Y \mid A = 0}} - \paren{\E{C \mid A = 1} - \E{C \mid A = 0}}
&& \text{(equi-confounding)}
\end{align}

The condition $0 < \Pr[A = 1] < 1$ makes every conditional mean well defined.
:::

::: {#exm-aspirin-blood-pressure}
## Aspirin and Blood Pressure

Suppose unmeasured $U$ (e.g., history of heart disease) confounds the effect of aspirin $A$ on blood pressure $Y$,
and we also measured blood pressure just before treatment, $C$, a negative outcome control (Figure 7.13).
@eq-difference-in-differences subtracts the treated-untreated difference in $C$ (pure confounding) from the treated-untreated difference in $Y$ (effect plus confounding).
The arrow $C \rightarrow Y$ is not necessary for $C$ to be a negative outcome control.
:::

::: {#rem-proximal-inference}
## Limits of Difference-in-Differences, and Proximal Inference

Difference-in-differences (Card 1990; Meyer 1995;
Angrist and Krueger 1999) is a somewhat restrictive use of negative outcome controls:
it needs $Y$ and $C$ on the same scale and additive equi-confounding;
Sofer et al. (2016) describe more general methods.

With both a negative outcome control $C$ and a negative treatment control $Z$,
the effect can be identified nonparametrically despite unmeasured $U$ under additional assumptions;
for discrete $U$, $C$, $Z$ with $C$ and $Z$ having at least as many levels as $U$, identification holds quite generally (Miao et al. 2018).
This is **proximal causal inference** (Cui et al. 2024);
Figure 7.15 (not drawn here) is an example [@hernan2020causal, p. 106].
:::

---

::: {.callout-note title="Technical Point 7.4: The Front Door Criterion"}
When an unmeasured $U$ blocks standardization and IP weighting, a mediator can sometimes stand in.
@prp-front-door gives Pearl's (1995) front door formula and its proof [@hernan2020causal, p. 107].
(Pearl's "backdoor formula" is what the book calls standardization or the point-treatment g-formula, @thm-standardization-ch7.)
:::

::: {#prp-front-door}
## Front Door Formula

Let $A$ be a treatment, $Y$ a binary outcome, and $M$ a discrete mediator, related as in Figure 7.14:
an unmeasured $U$ causes $A$ and $Y$,
$M$ fully mediates the effect of $A$ on $Y$,
and $M$ shares no unmeasured causes with $A$ or $Y$.
Assume an FFRCISTG model for Figure 7.14,
well-defined counterfactuals $Y^m$ under interventions on $M$,
consistency,
and positivity: $\Pr[A = a'] > 0$ and $\Pr[M = m \mid A = a'] > 0$ for every $a'$ and every $m$ with $\Pr[M = m \mid A = a] > 0$.
Then

$$
\Pr[Y^a = 1] = \sum_m \Pr[M = m \mid A = a] \sum_{a'} \Pr[Y = 1 \mid M = m, A = a'] \Pr[A = a']
$$ {#eq-front-door}
:::

::: {.proof}
1. By the law of total probability, $\Pr[Y^a = 1] = \sum_m \Pr[M^a = m] \Pr[Y^a = 1 \mid M^a = m]$.
2. $\Pr[M^a = m] = \Pr[M = m \mid A = a]$, because there is no confounding for the effect of $A$ on $M$ ($M^a \ind A$), and by consistency.
3. $\Pr[Y^a = 1 \mid M^a = m] = \Pr[Y^m = 1]$, because
   (i) $Y^a = Y^m$ when $M^a = m$ ($A$ affects $Y$ only through $M$), and
   (ii) $Y^m \ind M^a$ by d-separation on the SWIG for the joint intervention setting $M$ to $m$ and $A$ to $a$.
4. $\Pr[Y^m = 1] = \sum_{a'} \Pr[Y = 1 \mid M = m, A = a'] \Pr[A = a']$, by conditional exchangeability $Y^m \ind M \mid A$ on the SWIG intervening on $M$ alone, positivity, and consistency (standardization over $A$).
5. Substituting steps 2-4 into step 1 gives @eq-front-door.

This proof requires well-defined counterfactuals $Y^m$;
Technical Points 21.11 and 21.12 give proofs without that condition [@hernan2020causal, p. 107].
:::

## Summary

---

1. **Confounding** is the bias due to common causes of treatment and outcome, which create open backdoor paths.
2. **No confounding is equivalent to marginal exchangeability** (under faithfulness);
   with confounding, a sufficient set $L$ of measured non-descendants of $A$ gives conditional exchangeability, and standardization or IP weighting identifies the average causal effect.
3. Under faithfulness and an FFRCISTG model, $Y^a \ind A \mid L$ holds **if and only if** $L$ satisfies the **backdoor criterion**.
4. Adjusting for a collider such as $L$ in Figure 7.4 (M-bias) creates selection bias;
   in Figure 7.5 no adjustment set based on $L$ alone works.
5. The **traditional** associational definition of confounder can recommend harmful adjustment;
   confounding is absolute, but "confounder" is relative to the other adjustment variables.
6. **SWIGs** split the treatment node and show exchangeability directly via d-separation.
7. **G-methods** and conventional stratification-based methods both require conditional exchangeability;
   difference-in-differences, proximal inference, the front door criterion, and instrumental variables rely on other unverifiable assumptions.

::: {.notes}
**Looking ahead:**

- **Chapter 8**: selection bias, the other structural source of lack of exchangeability through open paths.
- **Chapter 9**: measurement bias.
- **Chapters 12-15**: IP weighting, standardization, g-estimation, and outcome regression in practice.
:::

## References

---

::: {#refs}
:::
